---
title: "Dimensionality Reduction Exploration"
author: "Juliana Pinto"
format: html
editor: visual
params:
  dataset_id: "GSExxxxxx"   # ← EDIT THIS LINE before running in RStudio
---

<!--
================================================================================
HOW TO USE THIS NOTEBOOK

 1. Copy this template to a dataset-specific file:
      cp 00_dimred_template.qmd dimred_exploration_<DATASET>.qmd
    (the copy is gitignored by the **/*_GSE*.qmd rule).

 2. Open the copy in RStudio — ideally by double-clicking
    deconv_gastric_cancer.Rproj first, so the working directory is the
    repo root and here::here() resolves correctly.

 3. Edit the YAML header above:
      params:
        dataset_id: "GSE246662"   # your dataset

 4. Run cell-by-cell (Ctrl/Cmd+Enter) or render the whole notebook
    with the green "Render" button in RStudio.

 Alternative: render from the terminal without editing the YAML:
      quarto render 00_dimred_template.qmd \
        -P dataset_id:GSE246662 \
        --output dimred_exploration_GSE246662.html

 This notebook expects that script 07a_dimred.R has already been run for
 the chosen dataset. It reads pca_seurat_path and uses dims_to_use/n_pcs
 from the config. After inspection, update dims_to_use in the config
 and only re-run 07b_clustering.R (PCA does not need to be recomputed).
================================================================================
-->

# Dimensionality Reduction: `r params$dataset_id`

Inspects the output of `07a_dimred.R` so an informed number of PCs can be set in `dims_to_use` before running `07b_clustering.R`.

## Setup

```{r libs, message=FALSE, warning=FALSE}
library(Seurat)
library(ggplot2)
library(patchwork)

theme_set(theme_classic())
```

```{r load-config}
# Loads config_<DATASET>.R using params$dataset_id. Paths resolve via here::here(),
# which requires this notebook to be run from the repo root (RStudio opens the
# project at the root automatically when you launch via the .Rproj file).
config_path <- file.path(
  here::here("chapter_2_sc_deconv_benchmarking", "01_sc_matrix",
             "scripts", "01_sc_pre_proc", "configs"),
  paste0("config_", params$dataset_id, ".R")
)
source(config_path)
```

```{r helper-function}
# Helper: map Ensembl IDs to gene symbols
ens_to_label <- function(ensembl_ids, seu) {
  if ("gene_symbol" %in% colnames(seu[["RNA"]]@meta.data)) {
    gene_meta <- seu[["RNA"]]@meta.data
    symbol_lookup <- setNames(gene_meta$gene_symbol, rownames(seu))
    labels <- symbol_lookup[ensembl_ids]
    labels[is.na(labels) | labels == ""] <- ensembl_ids[is.na(labels) | labels == ""]
    return(unname(labels))
  }
  return(ensembl_ids)
}
```

```{r load-object}
seu <- readRDS(pca_seurat_path)

cat("Dataset:   ", params$dataset_id, "\n")
cat("Dimensions:", nrow(seu), "genes x", ncol(seu), "cells\n")
cat("Samples:   ", length(unique(seu$orig.ident)), "\n")
cat("HVGs:      ", length(VariableFeatures(seu)), "\n")
```

## 1. ElbowPlot - Selecting PCs

```{r elbow-plot, fig.width=10, fig.height=5}
stdev <- seu[["pca"]]@stdev
var_explained <- (stdev^2 / sum(stdev^2)) * 100

p1 <- ElbowPlot(seu, ndims = n_pcs, reduction = "pca") +
  geom_vline(xintercept = max(dims_to_use), linetype = "dashed", color = "red") +
  ggtitle(paste("ElbowPlot - cutoff at PC", max(dims_to_use)))

# Cumulative variance
cum_var <- data.frame(
  PC = 1:length(var_explained),
  cumulative = cumsum(var_explained)
)

p2 <- ggplot(cum_var, aes(x = PC, y = cumulative)) +
  geom_line() +
  geom_point(size = 0.8) +
  geom_vline(xintercept = max(dims_to_use), linetype = "dashed", color = "red") +
  ggtitle("Cumulative variance explained") +
  ylab("Cumulative % variance") +
  xlab("PC")

p1 + p2
```

```{r variance-summary}
cat("Variance explained by PC1:", round(var_explained[1], 1), "%\n")
cat("Variance explained by PC1-PC", max(dims_to_use), ": ",
    round(sum(var_explained[dims_to_use]), 1), "%\n")
cat("Variance explained by PC1-PC", n_pcs, ": ",
    round(sum(var_explained), 1), "%\n")
```

```{r pc-recommendations}
# How many PCs are needed to reach common variance thresholds?
cum_var_vec <- cumsum(var_explained)
n_pc_50 <- which(cum_var_vec >= 50)[1]
n_pc_75 <- which(cum_var_vec >= 75)[1]
n_pc_90 <- which(cum_var_vec >= 90)[1]

cat("PCs needed for 50% variance:", n_pc_50, "\n")
cat("PCs needed for 75% variance:", n_pc_75, "\n")
cat("PCs needed for 90% variance:", n_pc_90, "\n")
```

## 2. PCA Loadings - Top genes per PC

Top 5 positive and top 5 negative loadings per PC (PC1-PC4). Helpful to tell whether early PCs are driven by biology (immune vs stromal vs epithelial markers) or by technical noise / single genes.

```{r pca-loadings, fig.width=12, fig.height=10}
plots_pca <- list()
for (pc in 1:4) {
  loadings <- seu[["pca"]]@feature.loadings[, pc]
  top_pos <- names(sort(loadings, decreasing = TRUE))[1:5]
  top_neg <- names(sort(loadings, decreasing = FALSE))[1:5]
  top_genes <- c(top_pos, top_neg)
  top_labels <- ens_to_label(top_genes, seu)

  df <- data.frame(
    gene = factor(top_labels, levels = rev(top_labels)),
    loading = loadings[top_genes]
  )

  plots_pca[[pc]] <- ggplot(df, aes(x = loading, y = gene, fill = loading > 0)) +
    geom_col() +
    scale_fill_manual(values = c("steelblue", "firebrick")) +
    ggtitle(paste0("PC", pc)) +
    theme(legend.position = "none") +
    xlab("Loading") + ylab(NULL)
}

wrap_plots(plots_pca, ncol = 2)
```

## 3. Variance explained per PC

```{r var-per-pc, fig.width=10, fig.height=5}
df_var <- data.frame(
  PC = 1:length(var_explained),
  variance = var_explained
)

ggplot(df_var[1:30, ], aes(x = PC, y = variance)) +
  geom_col(fill = ifelse(1:30 <= max(dims_to_use), "steelblue", "grey70")) +
  geom_vline(xintercept = max(dims_to_use) + 0.5, linetype = "dashed", color = "red") +
  ggtitle("Variance explained per PC (first 30)") +
  ylab("% variance") +
  xlab("PC")
```

## 4. Summary

```{r summary}
cat("Dataset:       ", params$dataset_id, "\n")
cat("Cells:         ", ncol(seu), "\n")
cat("Genes:         ", nrow(seu), "\n")
cat("HVGs:          ", length(VariableFeatures(seu)), "\n")
cat("PCs computed:  ", n_pcs, "\n")
cat("Current dims_to_use:", min(dims_to_use), "-", max(dims_to_use), "\n")
cat("Variance explained (selected PCs):",
    round(sum(var_explained[dims_to_use]), 1), "%\n")
```

## 5. Conclusions

Notes (fill in while inspecting the plots above):

-   **ElbowPlot**: Is the PC cutoff at `r max(dims_to_use)` appropriate?
    -   Look for where the curve flattens (the "elbow")
    -   Current cutoff captures `r round(sum(var_explained[dims_to_use]), 1)`% of variance
-   **PCA Loadings**: Do the top genes per PC make biological sense?
    -   Expected: immune genes vs stromal genes vs epithelial genes on different PCs
    -   Single-gene-dominated PCs may indicate technical artifacts (mitochondrial, ribosomal, or HB contamination)
-   **Action**: If `dims_to_use` needs adjustment, update the config and re-run only `07b_clustering.R` (no need to recompute PCA)
