---
title: "QC Exploration"
author: "Juliana Pinto"
format: html
editor: visual
params:
  dataset_id: "GSExxxxxx"   # ← EDIT THIS LINE before running in RStudio
---

<!--
================================================================================
HOW TO USE THIS NOTEBOOK

 1. Copy this template to a dataset-specific file:
      cp 00_qc_template.qmd QC_exploration_<DATASET>.qmd
    (the copy is gitignored by the **/*_GSE*.qmd rule).

 2. Open the copy in RStudio — ideally by double-clicking
    deconv_gastric_cancer.Rproj first, so the working directory is the
    repo root and here::here() resolves correctly.

 3. Edit the YAML header above:
      params:
        dataset_id: "GSE246662"   # your dataset

 4. Run cell-by-cell (Ctrl/Cmd+Enter) or render the whole notebook
    with the green "Render" button in RStudio.

 Alternative: render from the terminal without editing the YAML:
      quarto render 00_qc_template.qmd \
        -P dataset_id:GSE246662 \
        --output QC_exploration_GSE246662.html

 This notebook expects that script 03_qc_metrics.R has already been run
 for the chosen dataset. It only reads the per-sample RDS files written
 by that script.
================================================================================
-->

# QC Exploration: `r params$dataset_id`

Exploratory inspection of the QC metrics computed by `03_qc_metrics.R`, **before** filtering. Use this notebook to decide on the cutoffs to set in `config_<DATASET>.R` before running `04_qc_filtering.R`.

## Setup

```{r libs, message=FALSE, warning=FALSE}
library(Seurat)
library(ggplot2)
library(patchwork)

theme_set(theme_classic())
```

```{r load-config}
# Loads config_<DATASET>.R using params$dataset_id. Paths resolve via here::here(),
# which requires this notebook to be run from the repo root (RStudio opens the
# project at the root automatically when you launch via the .Rproj file).
config_path <- file.path(
  here::here("chapter_2_sc_deconv_benchmarking", "01_sc_matrix",
             "scripts", "01_sc_pre_proc", "configs"),
  paste0("config_", params$dataset_id, ".R")
)
source(config_path)
```

```{r load-objects}
qc_files <- list.files(
  seurat_qc_metrics_dir,
  pattern = "_seurat_qc_metrics.rds$",
  full.names = TRUE
)

sample_names <- gsub("_seurat_qc_metrics.rds", "", basename(qc_files))

seu_list <- setNames(
  lapply(qc_files, readRDS),
  sample_names
)

cat("Samples loaded:", length(seu_list), "\n")
cat("Names:", paste(names(seu_list), collapse = ", "), "\n")
```

## 1. Overview

```{r summary-table}
qc_summary <- do.call(rbind, lapply(names(seu_list), function(s) {

  seu <- seu_list[[s]]

  data.frame(
    sample = s,
    cells = ncol(seu),
    median_features = median(seu$nFeature_RNA),
    median_counts = median(seu$nCount_RNA),
    median_mt = round(median(seu$percent.mt), 2),
    median_ribo = round(median(seu$percent.ribo), 2),
    median_hb = round(median(seu$percent.hb), 2),
    pct_doublets = if ("scDblFinder.class" %in% colnames(seu@meta.data)) {
      round(100 * mean(seu$scDblFinder.class == "doublet"), 1)
    } else {
      NA
    }
  )
}))

knitr::kable(qc_summary, caption = paste("QC summary -", params$dataset_id))
```

```{r total-cells}
cat("Total cells across all samples:", sum(qc_summary$cells), "\n")
```

## 2. Comparative boxplots (all samples)

Combine all metadata into a single dataframe for cross-sample comparison.

```{r combined-metadata}
combined_meta <- do.call(rbind, lapply(names(seu_list), function(s) {
  df <- seu_list[[s]]@meta.data
  df$sample <- s
  df
}))

combined_meta$sample <- factor(combined_meta$sample, levels = names(seu_list))
```

```{r boxplot-features-counts, fig.width=12, fig.height=5}
p1 <- ggplot(combined_meta, aes(x = sample, y = nFeature_RNA, fill = sample)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  ggtitle("nFeature_RNA per sample")

p2 <- ggplot(combined_meta, aes(x = sample, y = nCount_RNA, fill = sample)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  ggtitle("nCount_RNA per sample")

p1 + p2
```

```{r boxplot-percentages, fig.width=14, fig.height=5}
p1 <- ggplot(combined_meta, aes(x = sample, y = percent.mt, fill = sample)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.3) +
  geom_hline(yintercept = qc_max_percent_mt, linetype = "dashed", color = "red") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  ggtitle("percent.mt per sample")

p2 <- ggplot(combined_meta, aes(x = sample, y = percent.ribo, fill = sample)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  ggtitle("percent.ribo per sample")

p3 <- ggplot(combined_meta, aes(x = sample, y = percent.hb, fill = sample)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  ggtitle("percent.hb per sample")

p1 + p2 + p3
```

## 3. Violin plots per sample

```{r violin-main, fig.width=10, fig.height=5, results='asis'}
for (s in names(seu_list)) {

  cat("\n\n###", s, "\n\n")

  seu <- seu_list[[s]]

  print(
    VlnPlot(seu, features = c("nFeature_RNA", "nCount_RNA"), pt.size = 0) +
      plot_annotation(title = s)
  )
}
```

```{r violin-percentages, fig.width=10, fig.height=5, results='asis'}
for (s in names(seu_list)) {

  cat("\n\n###", s, "\n\n")

  seu <- seu_list[[s]]

  print(
    VlnPlot(seu, features = c("percent.mt", "percent.ribo", "percent.hb"), pt.size = 0) +
      plot_annotation(title = s)
  )
}
```

## 4. Histograms

```{r hist-features, fig.width=10, fig.height=4, results='asis'}
for (s in names(seu_list)) {

  cat("\n\n###", s, "\n\n")

  df <- seu_list[[s]]@meta.data

  p1 <- ggplot(df, aes(x = nFeature_RNA)) +
    geom_histogram(bins = 80, fill = "steelblue", color = "white") +
    geom_vline(xintercept = qc_min_features, linetype = "dashed", color = "red") +
    ggtitle(paste("nFeature_RNA -", s))

  p2 <- ggplot(df, aes(x = nCount_RNA)) +
    geom_histogram(bins = 80, fill = "steelblue", color = "white") +
    geom_vline(xintercept = qc_min_counts, linetype = "dashed", color = "red") +
    ggtitle(paste("nCount_RNA -", s))

  print(p1 + p2)
}
```

```{r hist-mt-density, fig.width=10, fig.height=4, results='asis'}
for (s in names(seu_list)) {

  cat("\n\n###", s, "\n\n")

  df <- seu_list[[s]]@meta.data

  p1 <- ggplot(df, aes(x = percent.mt)) +
    geom_histogram(bins = 80, fill = "steelblue", color = "white") +
    geom_vline(xintercept = qc_max_percent_mt, linetype = "dashed", color = "red") +
    ggtitle(paste("percent.mt -", s))

  p2 <- ggplot(df, aes(x = percent.mt)) +
    geom_density(fill = "steelblue", alpha = 0.3) +
    geom_vline(xintercept = qc_max_percent_mt, linetype = "dashed", color = "red") +
    geom_vline(xintercept = quantile(df$percent.mt, 0.99),
               linetype = "dotted", color = "darkred") +
    ggtitle(paste("percent.mt density -", s, "(dotted = p99)"))

  print(p1 + p2)
}
```

## 5. Scatter plots

```{r scatter-counts-features, fig.width=10, fig.height=5, results='asis'}
for (s in names(seu_list)) {

  cat("\n\n###", s, "\n\n")

  df <- seu_list[[s]]@meta.data

  p1 <- ggplot(df, aes(x = nCount_RNA, y = nFeature_RNA, color = percent.mt)) +
    geom_point(alpha = 0.3, size = 0.5) +
    scale_color_viridis_c() +
    ggtitle(paste("Counts vs Features -", s))

  p2 <- ggplot(df, aes(x = nCount_RNA, y = percent.mt)) +
    geom_point(alpha = 0.3, size = 0.5) +
    geom_hline(yintercept = qc_max_percent_mt, linetype = "dashed", color = "red") +
    ggtitle(paste("Counts vs percent.mt -", s))

  print(p1 + p2)
}
```

```{r scatter-counts-ribo, fig.width=10, fig.height=5, results='asis'}
for (s in names(seu_list)) {

  cat("\n\n###", s, "\n\n")

  df <- seu_list[[s]]@meta.data

  p1 <- ggplot(df, aes(x = nCount_RNA, y = percent.ribo)) +
    geom_point(alpha = 0.3, size = 0.5) +
    ggtitle(paste("Counts vs percent.ribo -", s))

  p2 <- ggplot(df, aes(x = nCount_RNA, y = percent.hb)) +
    geom_point(alpha = 0.3, size = 0.5) +
    ggtitle(paste("Counts vs percent.hb -", s))

  print(p1 + p2)
}
```

## 6. Doublet detection

```{r doublet-table, results='asis'}
for (s in names(seu_list)) {

  cat("\n\n###", s, "\n\n")

  seu <- seu_list[[s]]

  if ("scDblFinder.class" %in% colnames(seu@meta.data)) {
    tbl <- table(seu$scDblFinder.class)
    pct <- round(100 * tbl["doublet"] / sum(tbl), 1)
    cat("Singlet:", tbl["singlet"], "| Doublet:", tbl["doublet"],
        "| % Doublet:", pct, "%\n\n")
  } else {
    cat("Doublet detection not available\n\n")
  }
}
```

```{r doublet-barplot, fig.width=8, fig.height=4}
if ("scDblFinder.class" %in% colnames(combined_meta)) {

  doublet_summary <- do.call(rbind, lapply(names(seu_list), function(s) {
    seu <- seu_list[[s]]
    data.frame(
      sample = s,
      pct_doublet = 100 * mean(seu$scDblFinder.class == "doublet")
    )
  }))

  ggplot(doublet_summary, aes(x = sample, y = pct_doublet, fill = sample)) +
    geom_col() +
    geom_hline(yintercept = 5, linetype = "dashed", color = "red") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
    ylab("% Doublets") +
    ggtitle("Doublet rate per sample (dashed = 5% reference)")
}
```

```{r doublet-violin, fig.width=10, fig.height=4, results='asis'}
if ("doublet_score" %in% colnames(combined_meta)) {

  for (s in names(seu_list)) {

    cat("\n\n###", s, "\n\n")

    print(
      VlnPlot(seu_list[[s]], features = "doublet_score", pt.size = 0) +
        ggtitle(s)
    )
  }
}
```

## 7. Correlations

```{r correlations, results='asis'}
for (s in names(seu_list)) {

  cat("\n\n###", s, "\n\n")

  df <- seu_list[[s]]@meta.data

  cor_matrix <- cor(
    df[, c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.ribo", "percent.hb")],
    method = "spearman"
  )

  cat("```\n")
  print(round(cor_matrix, 3))
  cat("```\n")
}
```

## 8. Conclusions

Notes on sample quality and decisions (fill in while inspecting the plots above):

-   **Samples to exclude (biological):** *(e.g. metastases, non-primary tumor)*
-   **Samples to exclude (quality):** *(e.g. high MT%, low cell count)*
-   **Threshold adjustments needed:** *(based on distribution observations)*
    -   `qc_min_features`:
    -   `qc_max_features`:
    -   `qc_min_counts`:
    -   `qc_max_percent_mt`:
-   **Doublet removal strategy:**
