---
title: "Normalization & HVG Exploration"
author: "Juliana Pinto"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
params:
  dataset_id: "GSExxxxxx"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 6
)
```

# Normalization: `r params$dataset_id`

Inspects the output of `06_normalize_hvg.R`: normalization quality, HVG selection, and a top-HVG sanity check with gene symbols.

## Setup

```{r libs}
library(Seurat)
library(ggplot2)
library(patchwork)

theme_set(theme_classic())
```

```{r load-config}
config_path <- file.path(
  here::here("chapter_2_sc_deconv_benchmarking", "01_sc_matrix", "scripts", "01_sc_pre_proc", "configs"),
  paste0("config_", params$dataset_id, ".R")
)
source(config_path)
```

```{r helper}
# Map Ensembl IDs to gene symbols using the feature metadata
ens_to_label <- function(ensembl_ids, seu) {
  if ("gene_symbol" %in% colnames(seu[["RNA"]]@meta.data)) {
    gene_meta <- seu[["RNA"]]@meta.data
    symbol_lookup <- setNames(gene_meta$gene_symbol, rownames(seu))
    labels <- symbol_lookup[ensembl_ids]
    labels[is.na(labels) | labels == ""] <- ensembl_ids[is.na(labels) | labels == ""]
    return(unname(labels))
  }
  ensembl_ids
}
```

```{r load-objects}
seu_merged <- readRDS(merged_seurat_path)
seu_norm   <- readRDS(norm_seurat_path)

cat("Merged:    ", ncol(seu_merged), "cells x", nrow(seu_merged), "genes\n")
cat("Normalized:", ncol(seu_norm),   "cells x", nrow(seu_norm),   "genes\n")
cat("HVGs:      ", length(VariableFeatures(seu_norm)), "\n")
```

## 1. Counts vs normalized values per cell

```{r count-distributions}
df <- data.frame(
  total_counts = colSums(LayerData(seu_merged, layer = "counts")),
  norm_counts  = colSums(LayerData(seu_norm,   layer = "data"))
)

p1 <- ggplot(df, aes(x = total_counts)) +
  geom_histogram(bins = 60, fill = "steelblue") + scale_x_log10() +
  ggtitle("Raw counts per cell (log10)")

p2 <- ggplot(df, aes(x = norm_counts)) +
  geom_histogram(bins = 60, fill = "darkorange") +
  ggtitle("Normalized values per cell")

p1 + p2
```

## 2. Mean-variance / HVG plot

```{r hvg-plot}
top10 <- head(VariableFeatures(seu_norm), 10)
top10_labels <- ens_to_label(top10, seu_norm)

vfp <- VariableFeaturePlot(seu_norm)
LabelPoints(plot = vfp, points = top10, labels = top10_labels, repel = TRUE)
```

## 3. Top HVGs (with gene symbols)

```{r top-hvgs}
hvgs <- VariableFeatures(seu_norm)
hvg_table <- data.frame(
  rank   = seq_along(head(hvgs, 30)),
  ens_id = head(hvgs, 30),
  symbol = ens_to_label(head(hvgs, 30), seu_norm)
)
knitr::kable(hvg_table, caption = "Top 30 HVGs")
```

## 4. Notes / decisions

- Normalization method:
- Number of HVGs:
- Genes that were unexpected / interesting:
- Anything to revise before PCA:

---

*Render: `quarto render 00_norm_template.qmd -P dataset_id:GSExxxxxx --output norm_exploration_GSExxxxxx.html`*
