---
title: "Normalization & HVG Exploration"
author: "Juliana Pinto"
format: html
editor: visual
params:
  dataset_id: "GSExxxxxx"   # ← EDIT THIS LINE before running in RStudio
---

<!--
================================================================================
HOW TO USE THIS NOTEBOOK

 1. Copy this template to a dataset-specific file:
      cp 00_norm_template.qmd norm_exploration_<DATASET>.qmd
    (the copy is gitignored by the **/*_GSE*.qmd rule).

 2. Open the copy in RStudio — ideally by double-clicking
    deconv_gastric_cancer.Rproj first, so the working directory is the
    repo root and here::here() resolves correctly.

 3. Edit the YAML header above:
      params:
        dataset_id: "GSE246662"   # your dataset

 4. Run cell-by-cell (Ctrl/Cmd+Enter) or render the whole notebook
    with the green "Render" button in RStudio.

 Alternative: render from the terminal without editing the YAML:
      quarto render 00_norm_template.qmd \
        -P dataset_id:GSE246662 \
        --output norm_exploration_GSE246662.html

 This notebook expects that scripts 05_merge.R and 06_normalize_hvg.R
 have already been run for the chosen dataset. It reads:
   - merged_seurat_path (pre-normalization)
   - norm_seurat_path   (post-normalization + HVG)
================================================================================
-->

# Normalization & HVG: `r params$dataset_id`

Exploratory inspection of the normalization and HVG selection produced by `06_normalize_hvg.R`. Use this notebook to validate the normalization strategy (RC, scale factor, no log) and to check whether the selected HVGs make biological sense.

## Setup

```{r libs, message=FALSE, warning=FALSE}
library(Seurat)
library(ggplot2)
library(patchwork)

theme_set(theme_classic())
```

```{r load-config}
# Loads config_<DATASET>.R using params$dataset_id. Paths resolve via here::here(),
# which requires this notebook to be run from the repo root (RStudio opens the
# project at the root automatically when you launch via the .Rproj file).
config_path <- file.path(
  here::here("chapter_2_sc_deconv_benchmarking", "01_sc_matrix",
             "scripts", "01_sc_pre_proc", "configs"),
  paste0("config_", params$dataset_id, ".R")
)
source(config_path)
```

```{r helper-function}
# Helper: map Ensembl IDs to gene symbols
# Returns symbol if available, otherwise keeps Ensembl ID
# Uses named vector lookup (feature metadata rownames are numeric, not Ensembl)
ens_to_label <- function(ensembl_ids, seu) {
  if ("gene_symbol" %in% colnames(seu[["RNA"]]@meta.data)) {
    gene_meta <- seu[["RNA"]]@meta.data
    symbol_lookup <- setNames(gene_meta$gene_symbol, rownames(seu))
    labels <- symbol_lookup[ensembl_ids]
    labels[is.na(labels) | labels == ""] <- ensembl_ids[is.na(labels) | labels == ""]
    return(unname(labels))
  }
  return(ensembl_ids)
}
```

```{r load-objects}
# Merged (pre-normalization)
seu_merged <- readRDS(merged_seurat_path)

# Normalized (post-normalization + HVG)
seu_norm <- readRDS(norm_seurat_path)

cat("Merged:    ", ncol(seu_merged), "cells x", nrow(seu_merged), "genes\n")
cat("Normalized:", ncol(seu_norm),  "cells x", nrow(seu_norm),  "genes\n")
cat("HVGs:      ", length(VariableFeatures(seu_norm)), "\n")
```

## 1. Count distribution before vs after normalization

```{r count-distribution, fig.width=12, fig.height=5}
set.seed(42)
cells_sample <- sample(colnames(seu_merged), min(500, ncol(seu_merged)))

# Raw and normalized counts for one representative cell
raw_counts  <- GetAssayData(seu_merged, layer = "counts")[, cells_sample[1]]
norm_counts <- GetAssayData(seu_norm,   layer = "data")[, cells_sample[1]]

p1 <- ggplot(data.frame(x = raw_counts[raw_counts > 0]), aes(x = x)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  ggtitle("Raw counts (1 cell)") +
  xlab("Counts")

p2 <- ggplot(data.frame(x = norm_counts[norm_counts > 0]), aes(x = x)) +
  geom_histogram(bins = 50, fill = "darkorange", color = "white") +
  ggtitle("Normalized counts (1 cell)") +
  xlab("Normalized expression")

p1 + p2
```

## 2. Library size distribution per sample

```{r library-size, fig.width=10, fig.height=5}
df_merged <- seu_merged@meta.data
df_merged$sample <- seu_merged$orig.ident

p1 <- ggplot(df_merged, aes(x = sample, y = nCount_RNA, fill = sample)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  ggtitle("Library size per sample (raw)")

# After normalization
norm_lib_size <- colSums(GetAssayData(seu_norm, layer = "data"))
df_norm <- data.frame(
  sample = seu_norm$orig.ident,
  norm_total = norm_lib_size
)

p2 <- ggplot(df_norm, aes(x = sample, y = norm_total, fill = sample)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  ggtitle("Total expression per sample (normalized)")

p1 + p2
```

```{r free-merged}
rm(seu_merged)
gc()
```

## 3. Highly Variable Genes

```{r hvg-plot, fig.width=12, fig.height=6}
top10 <- head(VariableFeatures(seu_norm), 10)
top10_labels <- ens_to_label(top10, seu_norm)

plot1 <- VariableFeaturePlot(seu_norm)

# Label with gene symbols
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE,
                     labels = top10_labels)

plot1 + plot2 + plot_annotation(title = paste("HVGs -", params$dataset_id))
```

### Top 20 HVGs

```{r top-hvgs}
top20 <- head(VariableFeatures(seu_norm), 20)
top20_labels <- ens_to_label(top20, seu_norm)

hvg_table <- data.frame(
  rank = 1:20,
  ensembl_id = top20,
  gene_symbol = top20_labels,
  mapped = ifelse(top20_labels != top20, "yes", "no")
)

knitr::kable(hvg_table, caption = "Top 20 Highly Variable Genes")
```

### HVG overlap with QC genes

Check whether mitochondrial, ribosomal or hemoglobin genes are among the HVGs — they usually should not be. Presence may indicate incomplete filtering (MT) or contamination (HB: erythrocytes).

```{r hvg-qc-check}
hvgs       <- VariableFeatures(seu_norm)
hvg_labels <- ens_to_label(hvgs, seu_norm)

mt_in_hvg   <- hvgs[grepl(mito_pattern, hvg_labels)]
ribo_in_hvg <- hvgs[grepl(ribo_pattern, hvg_labels)]
hb_in_hvg   <- hvgs[grepl(hb_pattern,   hvg_labels)]

cat("Mitochondrial genes in HVGs:", length(mt_in_hvg), "\n")
if (length(mt_in_hvg) > 0) cat("  ", ens_to_label(mt_in_hvg, seu_norm), "\n")

cat("Ribosomal genes in HVGs:   ", length(ribo_in_hvg), "\n")
if (length(ribo_in_hvg) > 0) cat("  ", ens_to_label(ribo_in_hvg, seu_norm), "\n")

cat("Hemoglobin genes in HVGs:  ", length(hb_in_hvg), "\n")
if (length(hb_in_hvg) > 0) cat("  ", ens_to_label(hb_in_hvg, seu_norm), "\n")
```

## 4. Mean expression vs variance

```{r mean-var, fig.width=8, fig.height=6}
hvg_info <- HVFInfo(seu_norm)
hvg_info$is_hvg <- rownames(hvg_info) %in% VariableFeatures(seu_norm)

ggplot(hvg_info, aes(x = mean, y = variance, color = is_hvg)) +
  geom_point(size = 0.5, alpha = 0.3) +
  scale_color_manual(values = c("grey70", "red")) +
  scale_x_log10() +
  scale_y_log10() +
  ggtitle("Mean expression vs Variance") +
  labs(color = "HVG")
```

## 5. Expression of top HVGs across samples

```{r hvg-violin, fig.width=12, fig.height=8}
top6 <- head(VariableFeatures(seu_norm), 6)
top6_labels <- ens_to_label(top6, seu_norm)

plots <- list()
for (i in seq_along(top6)) {
  plots[[i]] <- VlnPlot(seu_norm, features = top6[i], pt.size = 0) +
    ggtitle(paste0(top6_labels[i], " (", top6[i], ")"))
}

wrap_plots(plots, ncol = 3)
```

## 6. Summary

```{r summary}
cat("Dataset:       ", params$dataset_id, "\n")
cat("Cells:         ", ncol(seu_norm), "\n")
cat("Genes:         ", nrow(seu_norm), "\n")
cat("Normalization: ", norm_method, "/ scale factor:", scale_factor, "\n")
cat("HVG method:    ", hvg_method, "/ n:", n_hvg, "\n")
cat("HVGs identified:", length(VariableFeatures(seu_norm)), "\n")

# Gene symbol mapping stats
if ("gene_symbol" %in% colnames(seu_norm[["RNA"]]@meta.data)) {
  gs <- seu_norm[["RNA"]]@meta.data$gene_symbol
  cat("Gene symbols mapped:  ", sum(!is.na(gs) & gs != ""), "/", length(gs), "\n")
  cat("Unmapped (Ensembl only):", sum(is.na(gs) | gs == ""), "\n")
}
```

## 7. Conclusions

Notes (fill in while inspecting the plots above):

-   **HVGs look reasonable?** *(check if top genes make biological sense)*
-   **MT / Ribo / HB genes in HVGs?** *(flag if present — MT/Ribo = incomplete filtering; HB = erythrocyte contamination)*
-   **Adjust `n_hvg`?** *(2000 is standard, may need more/less depending on heterogeneity)*
-   **Unmapped genes:** some HVGs may show only Ensembl IDs (e.g. immunoglobulin loci not in the mapping table)
