---
title: "ROGUE Cluster Purity Exploration"
author: "Juliana Pinto"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
params:
  dataset_id: "GSExxxxxx"
  resolution: 0.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 7
)
```

# ROGUE: `r params$dataset_id`

Inspects the output of `08_rogue.R`. ROGUE measures cluster purity (1 = pure, < 0.5 = mixed). Used to detect clusters that may need re-clustering or that are biologically heterogeneous.

## Setup

```{r libs}
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
library(pheatmap)

theme_set(theme_classic())
```

```{r load-config}
config_path <- file.path(
  here::here("chapter_2_sc_deconv_benchmarking", "01_sc_matrix", "scripts", "01_sc_pre_proc", "configs"),
  paste0("config_", params$dataset_id, ".R")
)
source(config_path)
```

```{r load-data}
seu <- readRDS(clustered_seurat_path)
cluster_col <- paste0("RNA_snn_res.", params$resolution)
Idents(seu) <- cluster_col

rogue_csv <- file.path(rogue_output_dir,
                       paste0(params$dataset_id, "_rogue_scores.csv"))
rogue_df  <- read.csv(rogue_csv, row.names = 1, check.names = FALSE)

cat("ROGUE matrix:", nrow(rogue_df), "samples x", ncol(rogue_df), "clusters\n")
```

## 1. Score table (long format)

```{r long-table}
long <- rogue_df %>%
  tibble::rownames_to_column("sample") %>%
  tidyr::pivot_longer(-sample, names_to = "cluster", values_to = "rogue") %>%
  mutate(cluster = sub("^X", "", cluster))

knitr::kable(
  long %>% arrange(rogue) %>% head(20),
  caption = "20 lowest-purity (cluster, sample) combinations"
)
```

## 2. Boxplot per cluster

```{r boxplot}
ggplot(long, aes(x = cluster, y = rogue)) +
  geom_boxplot(fill = "steelblue", alpha = 0.5) +
  geom_jitter(width = 0.15, size = 1) +
  geom_hline(yintercept = c(0.5, 0.8), linetype = "dashed", color = "red") +
  ylim(0, 1) +
  ggtitle("ROGUE per cluster (across samples)") +
  ylab("ROGUE") + xlab("Cluster")
```

## 3. Heatmap (cluster × sample)

```{r heatmap}
mat <- as.matrix(rogue_df)
mat[is.na(mat)] <- 0
pheatmap(
  mat,
  cluster_rows = FALSE, cluster_cols = FALSE,
  display_numbers = TRUE, number_format = "%.2f",
  color = colorRampPalette(c("red", "white", "steelblue"))(100),
  main = "ROGUE (cluster x sample)"
)
```

## 4. ROGUE on UMAP

```{r umap-rogue}
mean_rogue <- colMeans(rogue_df, na.rm = TRUE)
names(mean_rogue) <- sub("^X", "", names(mean_rogue))

cluster_levels <- as.character(seu@meta.data[[cluster_col]])
seu$rogue_score <- unname(mean_rogue[cluster_levels])

FeaturePlot(seu, features = "rogue_score", order = TRUE) +
  scale_color_gradient(low = "red", high = "steelblue", name = "ROGUE") +
  ggtitle("Mean ROGUE per cluster on UMAP")
```

## 5. Low-purity investigation

```{r low-purity}
low_clusters <- names(mean_rogue)[mean_rogue < 0.6]
cat("Clusters with mean ROGUE < 0.6:",
    if (length(low_clusters)) paste(low_clusters, collapse = ", ") else "none",
    "\n")
```

## 6. Notes / decisions

- Pure clusters (ROGUE > 0.8):
- Moderate clusters (0.5–0.8):
- Mixed clusters needing attention:
- Re-cluster / merge candidates:

---

*Render: `quarto render 00_rogue_template.qmd -P dataset_id:GSExxxxxx -P resolution:0.5 --output rogue_exploration_GSExxxxxx.html`*
