---
title: "ROGUE Cluster Purity Exploration"
author: "Juliana Pinto"
format: html
editor: visual
params:
  dataset_id: "GSExxxxxx"   # ← EDIT THIS LINE before running in RStudio
---

<!--
================================================================================
HOW TO USE THIS NOTEBOOK

 1. Copy this template to a dataset-specific file:
      cp 00_rogue_template.qmd rogue_exploration_<DATASET>.qmd
    (the copy is gitignored by the **/*_GSE*.qmd rule).

 2. Open the copy in RStudio — ideally by double-clicking
    deconv_gastric_cancer.Rproj first, so the working directory is the
    repo root and here::here() resolves correctly.

 3. Edit the YAML header above:
      params:
        dataset_id: "GSE246662"   # your dataset

 4. Run cell-by-cell (Ctrl/Cmd+Enter) or render the whole notebook
    with the green "Render" button in RStudio.

 Alternative: render from the terminal without editing the YAML:
      quarto render 00_rogue_template.qmd \
        -P dataset_id:GSE246662 \
        --output rogue_exploration_GSE246662.html

 The resolution to inspect is taken directly from `rogue_resolution`
 in config_<DATASET>.R — no need to edit it here. Change it in the
 config and re-run 08_rogue.R if you want a different resolution.

 This notebook expects that script 08_rogue.R has already been run for
 the chosen dataset/resolution. It reads:
   - clustered_seurat_path
   - <rogue_output_dir>/<DATASET>_rogue_scores_res<rogue_resolution>.csv
================================================================================
-->

# ROGUE: `r params$dataset_id`

Inspects the output of `08_rogue.R`. ROGUE measures cluster purity (1 = pure, < 0.5 = mixed). Used to detect clusters that may need re-clustering or that are biologically heterogeneous.

## Setup

```{r libs, message=FALSE, warning=FALSE}
library(Seurat)
library(ggplot2)
library(patchwork)
library(ROGUE)
library(dplyr)
library(tidyr)

theme_set(theme_classic())
```

```{r load-config}
# Loads config_<DATASET>.R using params$dataset_id. Paths resolve via here::here(),
# which requires this notebook to be run from the repo root (RStudio opens the
# project at the root automatically when you launch via the .Rproj file).
config_path <- file.path(
  here::here("chapter_2_sc_deconv_benchmarking", "01_sc_matrix",
             "scripts", "01_sc_pre_proc", "configs"),
  paste0("config_", params$dataset_id, ".R")
)
source(config_path)
```

```{r helper-function}
# Helper: map Ensembl IDs to gene symbols
ens_to_label <- function(ensembl_ids, seu) {
  if ("gene_symbol" %in% colnames(seu[["RNA"]]@meta.data)) {
    gene_meta <- seu[["RNA"]]@meta.data
    symbol_lookup <- setNames(gene_meta$gene_symbol, rownames(seu))
    labels <- symbol_lookup[ensembl_ids]
    labels[is.na(labels) | labels == ""] <- ensembl_ids[is.na(labels) | labels == ""]
    return(unname(labels))
  }
  return(ensembl_ids)
}
```

## 1. Load data

```{r load-data}
seu <- readRDS(clustered_seurat_path)
cluster_col <- paste0("RNA_snn_res.", rogue_resolution)
Idents(seu) <- cluster_col

rogue_csv <- file.path(
  rogue_output_dir,
  paste0(params$dataset_id, "_rogue_scores_res", rogue_resolution, ".csv")
)
rogue_scores <- read.csv(rogue_csv, row.names = 1, check.names = FALSE)
rogue_mat <- as.matrix(rogue_scores)

cat("Dataset:   ", params$dataset_id, "\n")
cat("Cells:     ", ncol(seu), "\n")
cat("Resolution:", rogue_resolution, "\n")
cat("Clusters:  ", ncol(rogue_scores), "\n")
cat("Samples:   ", nrow(rogue_scores), "\n")
```

## 2. Score table (long format)

```{r long-table}
long <- rogue_scores %>%
  tibble::rownames_to_column("sample") %>%
  tidyr::pivot_longer(-sample, names_to = "cluster", values_to = "rogue") %>%
  mutate(cluster = sub("^X", "", cluster))

knitr::kable(
  long %>% arrange(rogue) %>% head(20),
  caption = "20 lowest-purity (cluster, sample) combinations"
)
```

```{r rogue-summary}
# Mean / min / max per cluster
cluster_means <- colMeans(rogue_mat, na.rm = TRUE)
cluster_summary <- data.frame(
  cluster    = names(cluster_means),
  mean_rogue = round(cluster_means, 3),
  min_rogue  = round(apply(rogue_mat, 2, min, na.rm = TRUE), 3),
  max_rogue  = round(apply(rogue_mat, 2, max, na.rm = TRUE), 3)
)
cluster_summary <- cluster_summary[order(cluster_summary$mean_rogue), ]

knitr::kable(cluster_summary, row.names = FALSE,
             caption = "ROGUE summary per cluster (sorted by mean)")

cat("\nOverall mean ROGUE:", round(mean(rogue_mat, na.rm = TRUE), 3), "\n")
```

## 3. Boxplot - ROGUE per cluster

```{r rogue-boxplot, fig.width=10, fig.height=6}
rogue_long <- as.data.frame(as.table(rogue_mat))
colnames(rogue_long) <- c("sample", "cluster", "rogue")
rogue_long$rogue <- as.numeric(rogue_long$rogue)
rogue_long <- rogue_long[!is.na(rogue_long$rogue), ]

ggplot(rogue_long, aes(x = cluster, y = rogue, fill = cluster)) +
  geom_boxplot(outlier.shape = 21) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "darkgreen", alpha = 0.6) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red",       alpha = 0.6) +
  annotate("text", x = 0.5, y = 0.82, label = "pure (>0.8)",  hjust = 0, size = 3, color = "darkgreen") +
  annotate("text", x = 0.5, y = 0.48, label = "mixed (<0.5)", hjust = 0, size = 3, color = "red") +
  ylim(0, 1) +
  ggtitle(paste("ROGUE per cluster -", params$dataset_id, "(res", rogue_resolution, ")")) +
  xlab("Cluster") + ylab("ROGUE score") +
  theme(legend.position = "none")
```

## 4. Heatmap - ROGUE cluster x sample

```{r rogue-heatmap, fig.width=12, fig.height=5}
ggplot(rogue_long, aes(x = cluster, y = sample, fill = rogue)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", rogue)), size = 3) +
  scale_fill_gradient2(low = "red", mid = "white", high = "steelblue",
                       midpoint = 0.5, limits = c(0, 1),
                       name = "ROGUE") +
  theme_minimal() +
  ggtitle(paste("ROGUE heatmap -", params$dataset_id, "(res", rogue_resolution, ")")) +
  xlab("Cluster") + ylab("Sample")
```

## 5. ROGUE built-in boxplot

```{r rogue-builtin-boxplot, fig.width=10, fig.height=6}
rogue.boxplot(rogue_scores)
```

## 6. ROGUE on UMAP

```{r rogue-umap, fig.width=14, fig.height=6}
# Map mean ROGUE per cluster to each cell
rogue_cluster_col <- paste0("RNA_snn_res.", rogue_resolution)

# Debug: what cluster labels look like in each source
cat("Cluster labels in Seurat (first 5):",
    head(unique(as.character(seu@meta.data[[rogue_cluster_col]]))), "\n")
cat("Cluster labels in ROGUE matrix:",
    colnames(rogue_mat), "\n")

# Build map — colnames(rogue_mat) may have "X" prefix if read.csv was called
# without check.names = FALSE (R prepends "X" to numeric names). Try direct
# match first; if it fails, strip "X" and retry.
cluster_rogue_map <- setNames(cluster_means, colnames(rogue_mat))

cell_clusters <- as.character(seu@meta.data[[rogue_cluster_col]])
rogue_values  <- cluster_rogue_map[cell_clusters]

if (all(is.na(rogue_values))) {
  cat("Direct match failed. Trying without 'X' prefix...\n")
  clean_names <- sub("^X", "", colnames(rogue_mat))
  cluster_rogue_map <- setNames(cluster_means, clean_names)
  rogue_values <- cluster_rogue_map[cell_clusters]
}

cat("Matched cells:", sum(!is.na(rogue_values)), "/", length(rogue_values), "\n")

seu$rogue_score <- unname(rogue_values)

p1 <- DimPlot(seu, group.by = rogue_cluster_col, label = TRUE) +
  ggtitle(paste("Clusters (res", rogue_resolution, ")")) +
  NoLegend()

# Only plot ROGUE on UMAP if values are not all NA
if (any(!is.na(seu$rogue_score))) {
  p2 <- FeaturePlot(seu, features = "rogue_score", order = TRUE) +
    scale_color_gradient2(low = "red", mid = "white", high = "steelblue",
                          midpoint = 0.5, limits = c(0, 1),
                          name = "ROGUE") +
    ggtitle("Mean ROGUE per cluster on UMAP")
  p1 + p2
} else {
  cat("WARNING: Could not map ROGUE scores to cells. Check cluster label formats.\n")
  p1
}
```

## 7. Low-purity clusters - investigation

```{r low-purity, fig.width=14, fig.height=10}
# Identify clusters with mean ROGUE < 0.5
low_purity <- cluster_summary$cluster[cluster_summary$mean_rogue < 0.5]

if (length(low_purity) > 0) {
  cat("Low purity clusters (mean ROGUE < 0.5):",
      paste(low_purity, collapse = ", "), "\n\n")

  # Check QC metrics for these clusters
  Idents(seu) <- rogue_cluster_col

  plots_lp <- list()
  for (feat in c("nCount_RNA", "nFeature_RNA", "percent.mt")) {
    if (feat %in% colnames(seu@meta.data)) {
      plots_lp[[feat]] <- VlnPlot(seu, features = feat, pt.size = 0) +
        ggtitle(feat) + NoLegend()
    }
  }

  if (length(plots_lp) > 0) wrap_plots(plots_lp, ncol = 2)
} else {
  cat("All clusters have mean ROGUE >= 0.5\n")
}
```

## 8. ROGUE per sample (global)

```{r rogue-per-sample}
sample_means <- rowMeans(rogue_mat, na.rm = TRUE)
sample_summary <- data.frame(
  sample     = names(sample_means),
  mean_rogue = round(sample_means, 3)
)
sample_summary <- sample_summary[order(sample_summary$mean_rogue), ]

knitr::kable(sample_summary, row.names = FALSE,
             caption = "Mean ROGUE per sample (across all clusters)")
```

## 9. Summary

```{r final-summary}
cat("Dataset:           ", params$dataset_id, "\n")
cat("Resolution:        ", rogue_resolution, "\n")
cat("Clusters:          ", ncol(rogue_mat), "\n")
cat("Samples:           ", nrow(rogue_mat), "\n")
cat("Overall mean ROGUE:", round(mean(rogue_mat, na.rm = TRUE), 3), "\n")
cat("\nClusters with mean ROGUE < 0.5:", length(low_purity), "\n")
if (length(low_purity) > 0) cat("  ->", paste(low_purity, collapse = ", "), "\n")
cat("Clusters with mean ROGUE > 0.8:", sum(cluster_means > 0.8), "\n")
```

## 10. Conclusions

Notes (fill in while inspecting the plots above):

-   **Overall ROGUE**: is it high (\>0.75)? Clustering is generally good.
-   **Low-purity clusters**: may contain mixed cell types or be over-split. Consider:
    -   Merging with neighboring clusters
    -   Re-clustering at a lower resolution
    -   These may resolve after integration
-   **Sample-specific variation**: if ROGUE varies a lot across samples for the same cluster, this may indicate batch effects
-   **Relation to batch effect**: low ROGUE in sample-dominated clusters is expected before integration
-   **Action**: use ROGUE together with marker genes (script 09) to decide on final annotation
