---
title: "Clustering Exploration"
author: "Juliana Pinto"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
params:
  dataset_id: "GSExxxxxx"
  resolution: 0.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 7
)
```

# Clustering: `r params$dataset_id`

Inspects the output of `07b_clustering.R`: UMAP, clusters, per-sample composition, QC per cluster, and known marker genes.

## Setup

```{r libs}
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)

theme_set(theme_classic())
```

```{r load-config}
config_path <- file.path(
  here::here("scripts", "sc_pre_proc", "configs"),
  paste0("config_", params$dataset_id, ".R")
)
source(config_path)
```

```{r helper}
ens_to_label <- function(ensembl_ids, seu) {
  if ("gene_symbol" %in% colnames(seu[["RNA"]]@meta.data)) {
    gene_meta <- seu[["RNA"]]@meta.data
    symbol_lookup <- setNames(gene_meta$gene_symbol, rownames(seu))
    labels <- symbol_lookup[ensembl_ids]
    labels[is.na(labels) | labels == ""] <- ensembl_ids[is.na(labels) | labels == ""]
    return(unname(labels))
  }
  ensembl_ids
}

# Looks up an Ensembl ID from a gene symbol
symbol_to_ens <- function(symbol, seu) {
  gene_meta <- seu[["RNA"]]@meta.data
  hits <- rownames(seu)[!is.na(gene_meta$gene_symbol) &
                        gene_meta$gene_symbol == symbol]
  if (length(hits) == 0) return(NA_character_)
  hits[1]
}

feature_plot_labeled <- function(seu, symbols) {
  ens_ids <- vapply(symbols, symbol_to_ens, character(1), seu = seu)
  valid <- !is.na(ens_ids)
  if (!any(valid)) return(NULL)
  p <- FeaturePlot(seu, features = ens_ids[valid], order = TRUE,
                   combine = FALSE)
  for (i in seq_along(p)) {
    p[[i]] <- p[[i]] + ggtitle(symbols[valid][i])
  }
  wrap_plots(p)
}
```

```{r load-clustered}
seu <- readRDS(clustered_seurat_path)
cluster_col <- paste0("RNA_snn_res.", params$resolution)
Idents(seu) <- cluster_col

cat("Cells:", ncol(seu), "\n")
cat("Clusters at res", params$resolution, ":",
    length(unique(seu@meta.data[[cluster_col]])), "\n")
cat("Samples:", paste(unique(seu$orig.ident), collapse = ", "), "\n")
```

## 1. UMAP

```{r umap}
p1 <- DimPlot(seu, group.by = cluster_col, label = TRUE) +
  ggtitle(paste("Clusters (res", params$resolution, ")")) + NoLegend()
p2 <- DimPlot(seu, group.by = "orig.ident") +
  ggtitle("Sample")
p1 + p2
```

## 2. Resolutions overview

```{r umap-multi-res, fig.height=10}
res_cols <- grep("^RNA_snn_res", colnames(seu@meta.data), value = TRUE)
plots <- lapply(res_cols, function(rc) {
  DimPlot(seu, group.by = rc, label = TRUE) +
    ggtitle(rc) + NoLegend()
})
wrap_plots(plots, ncol = 2)
```

## 3. Composition per sample

```{r composition}
df <- seu@meta.data %>%
  group_by(orig.ident, .data[[cluster_col]]) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(orig.ident) %>% mutate(pct = n / sum(n) * 100)

ggplot(df, aes(x = orig.ident, y = pct, fill = .data[[cluster_col]])) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Cluster composition per sample") +
  ylab("% cells") + xlab(NULL)
```

## 4. QC per cluster

```{r qc-per-cluster, fig.height=8}
metrics <- c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.ribo")
plots <- lapply(metrics, function(m) {
  VlnPlot(seu, features = m, group.by = cluster_col, pt.size = 0) +
    NoLegend() + ggtitle(m)
})
wrap_plots(plots, ncol = 2)
```

## 5. Known markers (edit per dataset)

```{r markers-epithelial}
feature_plot_labeled(seu, c("EPCAM", "KRT8", "KRT18"))
```

```{r markers-tcell}
feature_plot_labeled(seu, c("CD3D", "CD3E", "CD8A", "CD4"))
```

```{r markers-bcell}
feature_plot_labeled(seu, c("CD79A", "MS4A1", "JCHAIN"))
```

```{r markers-myeloid}
feature_plot_labeled(seu, c("LYZ", "CD68", "S100A8", "S100A9"))
```

```{r markers-fibroblast-endothelial}
feature_plot_labeled(seu, c("COL1A1", "DCN", "PECAM1", "VWF"))
```

## 6. Notes / decisions

- Working resolution:
- Clusters that look biologically distinct:
- Clusters that may need merging / splitting:
- Suspect clusters (low QC, doublets, mixed markers):

---

*Render: `quarto render 00_clustering_template.qmd -P dataset_id:GSExxxxxx -P resolution:0.5 --output clustering_exploration_GSExxxxxx.html`*
