---
title: "Dimensionality Reduction Exploration"
author: "Juliana Pinto"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
params:
  dataset_id: "GSExxxxxx"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 6
)
```

# Dimensionality Reduction: `r params$dataset_id`

Inspects the output of `07a_dimred.R` (PCA only) so that an informed number of PCs can be set in the config before running `07b_clustering.R`.

## Setup

```{r libs}
library(Seurat)
library(ggplot2)
library(patchwork)

theme_set(theme_classic())
```

```{r load-config}
config_path <- file.path(
  here::here("scripts", "sc_pre_proc", "configs"),
  paste0("config_", params$dataset_id, ".R")
)
source(config_path)
```

```{r helper}
ens_to_label <- function(ensembl_ids, seu) {
  if ("gene_symbol" %in% colnames(seu[["RNA"]]@meta.data)) {
    gene_meta <- seu[["RNA"]]@meta.data
    symbol_lookup <- setNames(gene_meta$gene_symbol, rownames(seu))
    labels <- symbol_lookup[ensembl_ids]
    labels[is.na(labels) | labels == ""] <- ensembl_ids[is.na(labels) | labels == ""]
    return(unname(labels))
  }
  ensembl_ids
}
```

```{r load-pca}
seu <- readRDS(pca_seurat_path)
cat("Cells:", ncol(seu), " Genes:", nrow(seu), "\n")
cat("Available reductions:", paste(Reductions(seu), collapse = ", "), "\n")
```

## 1. ElbowPlot

```{r elbow}
ElbowPlot(seu, ndims = 50) +
  ggtitle(paste("Elbow Plot -", params$dataset_id))
```

## 2. Cumulative variance

```{r cumvar}
stdev <- seu[["pca"]]@stdev
var_explained <- (stdev^2) / sum(stdev^2) * 100
cum_var <- cumsum(var_explained)

df <- data.frame(PC = seq_along(stdev), var = var_explained, cum = cum_var)

p1 <- ggplot(df, aes(x = PC, y = var)) +
  geom_col(fill = "steelblue") +
  ggtitle("% Variance per PC") + ylab("% variance")

p2 <- ggplot(df, aes(x = PC, y = cum)) +
  geom_line() + geom_point(size = 0.8) +
  geom_hline(yintercept = c(50, 75, 90), linetype = "dashed", color = "red") +
  ggtitle("Cumulative variance") + ylab("Cumulative %")

p1 + p2
```

```{r pc-recommendations}
n_pc_50 <- which(cum_var >= 50)[1]
n_pc_75 <- which(cum_var >= 75)[1]
n_pc_90 <- which(cum_var >= 90)[1]
cat("PCs needed for 50% variance:", n_pc_50, "\n")
cat("PCs needed for 75% variance:", n_pc_75, "\n")
cat("PCs needed for 90% variance:", n_pc_90, "\n")
```

## 3. Top loadings (with gene symbols)

```{r loadings, fig.height=12}
loadings <- Loadings(seu, reduction = "pca")
top_n <- 10
n_pcs_show <- 6

top_loadings <- lapply(1:n_pcs_show, function(i) {
  pc <- loadings[, i]
  top <- sort(abs(pc), decreasing = TRUE)[1:top_n]
  data.frame(
    PC     = paste0("PC", i),
    gene   = names(top),
    symbol = ens_to_label(names(top), seu),
    loading = pc[names(top)]
  )
})
top_loadings_df <- do.call(rbind, top_loadings)

ggplot(top_loadings_df, aes(x = reorder(symbol, loading), y = loading)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  facet_wrap(~ PC, scales = "free", ncol = 2) +
  xlab(NULL) + ylab("Loading") +
  ggtitle("Top loadings (gene symbols)")
```

## 4. Notes / decisions

- Number of PCs to use in `07b_clustering.R`:
- Biologically meaningful PCs (top loadings):
- PCs dominated by technical noise / single genes:

---

*Render: `quarto render 00_dimred_template.qmd -P dataset_id:GSExxxxxx --output dimred_exploration_GSExxxxxx.html`*
