---
title: "Clustering Exploration - GSE163558"
author: "Juliana Pinto"
format: html
editor: visual
---

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)

theme_set(theme_classic())
```

```{r load-config}
source("~/doutorado/deconv_gastric_cancer/scripts/sc_pre_proc/configs/config_GSE163558.R")
```

```{r helper-function}
# Helper: map Ensembl IDs to gene symbols
ens_to_label <- function(ensembl_ids, seu) {
  if ("gene_symbol" %in% colnames(seu[["RNA"]]@meta.data)) {
    gene_meta <- seu[["RNA"]]@meta.data
    symbol_lookup <- setNames(gene_meta$gene_symbol, rownames(seu))
    labels <- symbol_lookup[ensembl_ids]
    labels[is.na(labels) | labels == ""] <- ensembl_ids[is.na(labels) | labels == ""]
    return(unname(labels))
  }
  return(ensembl_ids)
}

# Helper: FeaturePlot with gene symbol titles
feature_plot_labeled <- function(seu, ensembl_ids, symbols, ...) {
  # Create individual plots with gene symbol titles
  plots <- list()
  for (i in seq_along(ensembl_ids)) {
    label <- ens_to_label(ensembl_ids[i], seu)
    plots[[i]] <- FeaturePlot(seu, features = ensembl_ids[i], order = TRUE, ...) +
      ggtitle(label)
  }
  wrap_plots(plots, ncol = 2)
}
```

```{r load-object}
seu <- readRDS(clustered_seurat_path)

cat("Dataset:", dataset_id, "\n")
cat("Dimensions:", nrow(seu), "genes x", ncol(seu), "cells\n")
cat("Samples:", length(unique(seu$orig.ident)), "\n")
cat("Reductions:", paste(names(seu@reductions), collapse = ", "), "\n")
cat("dims_to_use:", min(dims_to_use), "-", max(dims_to_use), "\n")
```

## 1. UMAP - Clusters at different resolutions

```{r umap-resolutions, fig.width=14, fig.height=10}
plots_res <- list()
for (res in cluster_resolutions) {
  col_name <- paste0("RNA_snn_res.", res)
  if (col_name %in% colnames(seu@meta.data)) {
    n_cl <- length(unique(seu@meta.data[[col_name]]))
    plots_res[[as.character(res)]] <- DimPlot(seu, group.by = col_name, label = TRUE) +
      ggtitle(paste0("res = ", res, " (", n_cl, " clusters)")) +
      NoLegend()
  }
}

wrap_plots(plots_res, ncol = 2)
```

## 2. UMAP - By sample

```{r umap-sample, fig.width=10, fig.height=6}
DimPlot(seu, group.by = "orig.ident") +
  ggtitle("UMAP by sample")
```

```{r umap-split-sample, fig.width=14, fig.height=5}
DimPlot(seu, group.by = "orig.ident", split.by = "orig.ident", ncol = 4) +
  ggtitle("UMAP split by sample") +
  NoLegend()
```

## 3. Cluster composition per sample

```{r cluster-composition, fig.width=12, fig.height=6}
mid_res <- cluster_resolutions[ceiling(length(cluster_resolutions) / 2)]
mid_col <- paste0("RNA_snn_res.", mid_res)

df_comp <- seu@meta.data %>%
  group_by(across(all_of(c("orig.ident", mid_col)))) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(orig.ident) %>%
  mutate(pct = n / sum(n) * 100)

colnames(df_comp)[2] <- "cluster"

p1 <- ggplot(df_comp, aes(x = orig.ident, y = pct, fill = cluster)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle(paste("Cluster composition per sample (res", mid_res, ")")) +
  ylab("% cells") + xlab(NULL)

p2 <- ggplot(df_comp, aes(x = cluster, y = pct, fill = orig.ident)) +
  geom_col() +
  ggtitle(paste("Sample composition per cluster (res", mid_res, ")")) +
  ylab("% cells") + xlab("Cluster")

p1 + p2
```

## 4. Cluster sizes

```{r cluster-sizes}
for (res in cluster_resolutions) {
  col_name <- paste0("RNA_snn_res.", res)
  if (col_name %in% colnames(seu@meta.data)) {
    counts <- sort(table(seu@meta.data[[col_name]]), decreasing = TRUE)
    cat("\nResolution", res, "- Cluster sizes:\n")
    for (cl in names(counts)) {
      cat(sprintf("  Cluster %s: %d cells (%.1f%%)\n", cl, counts[cl], 100 * counts[cl] / ncol(seu)))
    }
  }
}
```

## 5. QC metrics per cluster

```{r qc-per-cluster, fig.width=14, fig.height=10}
mid_col <- paste0("RNA_snn_res.", mid_res)
Idents(seu) <- mid_col

qc_features <- c("nCount_RNA", "nFeature_RNA")
if ("percent.mt" %in% colnames(seu@meta.data)) qc_features <- c(qc_features, "percent.mt")
if ("percent.ribo" %in% colnames(seu@meta.data)) qc_features <- c(qc_features, "percent.ribo")

plots_qc <- list()
for (feat in qc_features) {
  plots_qc[[feat]] <- VlnPlot(seu, features = feat, pt.size = 0) +
    ggtitle(feat) + NoLegend()
}

wrap_plots(plots_qc, ncol = 2)
```

## 6. Feature plots - Known marker genes

```{r known-markers-setup}
# Known gastric/immune marker genes (symbols)
known_markers <- list(
  "Epithelial"  = c("EPCAM", "KRT18", "KRT19", "CDH1"),
  "T cells"     = c("CD3D", "CD3E", "CD4", "CD8A"),
  "B cells"     = c("CD79A", "MS4A1", "CD19"),
  "Myeloid"     = c("CD14", "LYZ", "CD68", "ITGAM"),
  "Fibroblasts" = c("COL1A1", "DCN", "VIM"),
  "Endothelial" = c("PECAM1", "VWF", "CDH5")
)

# Build symbol -> Ensembl lookup
gene_meta <- seu[["RNA"]]@meta.data
symbol_to_ens <- setNames(rownames(seu), gene_meta$gene_symbol)
```

```{r known-markers-epithelial, fig.width=14, fig.height=10}
cell_type <- "Epithelial"
symbols <- known_markers[[cell_type]]
ensembl_ids <- symbol_to_ens[symbols]
found <- ensembl_ids[!is.na(ensembl_ids)]
cat(cell_type, "- found:", length(found), "/", length(symbols), "\n")

feature_plot_labeled(seu, found, symbols) +
  plot_annotation(title = cell_type)
```

```{r known-markers-tcells, fig.width=14, fig.height=10}
cell_type <- "T cells"
symbols <- known_markers[[cell_type]]
ensembl_ids <- symbol_to_ens[symbols]
found <- ensembl_ids[!is.na(ensembl_ids)]
cat(cell_type, "- found:", length(found), "/", length(symbols), "\n")

feature_plot_labeled(seu, found, symbols) +
  plot_annotation(title = cell_type)
```

```{r known-markers-bcells, fig.width=14, fig.height=8}
cell_type <- "B cells"
symbols <- known_markers[[cell_type]]
ensembl_ids <- symbol_to_ens[symbols]
found <- ensembl_ids[!is.na(ensembl_ids)]
cat(cell_type, "- found:", length(found), "/", length(symbols), "\n")

feature_plot_labeled(seu, found, symbols) +
  plot_annotation(title = cell_type)
```

```{r known-markers-myeloid, fig.width=14, fig.height=10}
cell_type <- "Myeloid"
symbols <- known_markers[[cell_type]]
ensembl_ids <- symbol_to_ens[symbols]
found <- ensembl_ids[!is.na(ensembl_ids)]
cat(cell_type, "- found:", length(found), "/", length(symbols), "\n")

feature_plot_labeled(seu, found, symbols) +
  plot_annotation(title = cell_type)
```

```{r known-markers-fibroblasts, fig.width=14, fig.height=8}
cell_type <- "Fibroblasts"
symbols <- known_markers[[cell_type]]
ensembl_ids <- symbol_to_ens[symbols]
found <- ensembl_ids[!is.na(ensembl_ids)]
cat(cell_type, "- found:", length(found), "/", length(symbols), "\n")

feature_plot_labeled(seu, found, symbols) +
  plot_annotation(title = cell_type)
```

```{r known-markers-endothelial, fig.width=14, fig.height=8}
cell_type <- "Endothelial"
symbols <- known_markers[[cell_type]]
ensembl_ids <- symbol_to_ens[symbols]
found <- ensembl_ids[!is.na(ensembl_ids)]
cat(cell_type, "- found:", length(found), "/", length(symbols), "\n")

feature_plot_labeled(seu, found, symbols) +
  plot_annotation(title = cell_type)
```

## 7. Resolution comparison

```{r resolution-tree, fig.width=10, fig.height=8}
if (length(cluster_resolutions) >= 2) {
  res_low <- cluster_resolutions[1]
  res_high <- cluster_resolutions[length(cluster_resolutions)]

  col_low <- paste0("RNA_snn_res.", res_low)
  col_high <- paste0("RNA_snn_res.", res_high)

  cross_tab <- table(
    low = seu@meta.data[[col_low]],
    high = seu@meta.data[[col_high]]
  )

  cat("Cross-tabulation: res", res_low, "vs res", res_high, "\n\n")
  print(cross_tab)

  cross_df <- as.data.frame(cross_tab)
  colnames(cross_df) <- c("low_res", "high_res", "n_cells")

  ggplot(cross_df, aes(x = high_res, y = low_res, fill = n_cells)) +
    geom_tile() +
    geom_text(aes(label = ifelse(n_cells > 0, n_cells, "")), size = 3) +
    scale_fill_gradient(low = "white", high = "steelblue") +
    ggtitle(paste("Cluster mapping: res", res_low, "->", res_high)) +
    xlab(paste("Clusters at res", res_high)) +
    ylab(paste("Clusters at res", res_low))
}
```

## 8. Summary

```{r summary}
stdev <- seu[["pca"]]@stdev
var_explained <- (stdev^2 / sum(stdev^2)) * 100

cat("Dataset:", dataset_id, "\n")
cat("Cells:", ncol(seu), "\n")
cat("Genes:", nrow(seu), "\n")
cat("Samples:", length(unique(seu$orig.ident)), "\n")
cat("PCs used:", min(dims_to_use), "-", max(dims_to_use), "\n")
cat("Variance explained (selected PCs):", round(sum(var_explained[dims_to_use]), 1), "%\n")
cat("\nClusters per resolution:\n")
for (res in cluster_resolutions) {
  col_name <- paste0("RNA_snn_res.", res)
  if (col_name %in% colnames(seu@meta.data)) {
    cat("  res", res, ":", length(unique(seu@meta.data[[col_name]])), "clusters\n")
  }
}
```

## 9. Conclusions

Notes:

-   **Batch effect**: Do samples mix well in UMAP? *(if not, integration is critical)*
-   **Cluster composition**: Are clusters driven by biology or by sample? *(sample-dominated clusters may indicate batch effects)*
-   **QC metrics**: Any cluster with unusually high %MT or low nFeature? *(may be low-quality cells that escaped filtering)*
-   **Known markers**: Do clusters show expected marker expression? *(helps validate clustering resolution)*
-   **Resolution choice**: Which resolution best separates known cell types without over-splitting?
-   **Action**: If `dims_to_use` or resolutions need adjustment, update the config and re-run only `07b_clustering.R`
