---
title: "Pre-Annotation Exploration"
author: "Juliana Pinto"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
params:
  dataset_id: "GSExxxxxx"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 7
)
```

# Pre-Annotation: `r params$dataset_id`

Compares the three pre-annotation methods (SingleR HPCA, SingleR Blueprint, CellTypist) on the output of scripts `09a` + `09b` + `09c`.

## Setup

```{r libs}
library(Seurat)
library(ggplot2)
library(patchwork)
library(pheatmap)
library(dplyr)

theme_set(theme_classic())
```

```{r load-config}
config_path <- file.path(
  here::here("chapter_2_sc_deconv_benchmarking", "01_sc_matrix", "scripts", "01_sc_pre_proc", "configs"),
  paste0("config_", params$dataset_id, ".R")
)
source(config_path)
```

```{r load-object}
seu <- readRDS(annotated_seurat_path)

cat("Dataset:", dataset_id, "\n")
cat("Dimensions:", nrow(seu), "genes x", ncol(seu), "cells\n")
cat("Samples:", length(unique(seu$orig.ident)), "\n")

annot_cols <- grep("singler|celltypist", colnames(seu@meta.data), value = TRUE)
cat("Annotation columns:", paste(annot_cols, collapse = ", "), "\n")
```

## 1. UMAP — all annotations

```{r umap-all, fig.width=12, fig.height=10}
annotation_cluster_col <- paste0("RNA_snn_res.", annotation_resolution)

plots <- list(
  DimPlot(seu, group.by = annotation_cluster_col, label = TRUE) +
    ggtitle(paste("Clusters (res", annotation_resolution, ")")) + NoLegend()
)
if ("singler_hpca" %in% colnames(seu@meta.data)) {
  plots[["hpca"]] <- DimPlot(seu, group.by = "singler_hpca",
                             label = TRUE, repel = TRUE) +
    ggtitle("SingleR - HPCA") + NoLegend()
}
if ("singler_blueprint" %in% colnames(seu@meta.data)) {
  plots[["bp"]] <- DimPlot(seu, group.by = "singler_blueprint",
                           label = TRUE, repel = TRUE) +
    ggtitle("SingleR - Blueprint") + NoLegend()
}
if ("celltypist_label" %in% colnames(seu@meta.data)) {
  plots[["ct"]] <- DimPlot(seu, group.by = "celltypist_label",
                           label = TRUE, repel = TRUE) +
    ggtitle("CellTypist") + NoLegend()
}
wrap_plots(plots, ncol = 2)
```

## 2. Annotation counts

```{r counts}
for (col in annot_cols[!grepl("conf_score", annot_cols)]) {
  cat("\n---", col, "---\n")
  print(sort(table(seu@meta.data[[col]]), decreasing = TRUE))
}
```

## 3. CellTypist confidence

```{r ct-confidence, fig.width=10, fig.height=10}
if ("celltypist_conf_score" %in% colnames(seu@meta.data)) {
  p1 <- FeaturePlot(seu, features = "celltypist_conf_score", order = TRUE) +
    scale_color_gradient(low = "red", high = "steelblue", name = "Score") +
    ggtitle("CellTypist confidence on UMAP")

  p2 <- ggplot(seu@meta.data, aes(x = celltypist_conf_score)) +
    geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
    geom_vline(xintercept = 0.5, linetype = "dashed", color = "red") +
    ggtitle("CellTypist confidence distribution") +
    xlab("Confidence score") + ylab("Cells")

  p1 / p2
}
```

## 4. Cluster-level consensus

```{r consensus}
consensus <- seu@meta.data %>%
  group_by(across(all_of(annotation_cluster_col))) %>%
  summarise(
    n_cells = n(),
    hpca = if ("singler_hpca" %in% colnames(seu@meta.data))
      names(sort(table(singler_hpca), decreasing = TRUE))[1] else NA,
    hpca_pct = if ("singler_hpca" %in% colnames(seu@meta.data))
      round(max(table(singler_hpca)) / n() * 100, 1) else NA,
    blueprint = if ("singler_blueprint" %in% colnames(seu@meta.data))
      names(sort(table(singler_blueprint), decreasing = TRUE))[1] else NA,
    blueprint_pct = if ("singler_blueprint" %in% colnames(seu@meta.data))
      round(max(table(singler_blueprint)) / n() * 100, 1) else NA,
    celltypist = if ("celltypist_label" %in% colnames(seu@meta.data))
      names(sort(table(celltypist_label), decreasing = TRUE))[1] else NA,
    celltypist_pct = if ("celltypist_label" %in% colnames(seu@meta.data))
      round(max(table(celltypist_label)) / n() * 100, 1) else NA,
    .groups = "drop"
  )

knitr::kable(consensus, caption = "Dominant annotation per cluster (% agreement)")
```

## 5. Confusion matrices

```{r confusion-hpca, fig.width=10, fig.height=8}
if ("singler_hpca" %in% colnames(seu@meta.data)) {
  ct <- table(Cluster = seu@meta.data[[annotation_cluster_col]],
              HPCA    = seu$singler_hpca)
  ct_prop <- prop.table(ct, margin = 1) * 100
  pheatmap(ct_prop, fontsize = 7, display_numbers = TRUE,
           number_format = "%.0f", cluster_rows = FALSE, cluster_cols = FALSE,
           color = colorRampPalette(c("white", "steelblue"))(100),
           main = "Cluster vs SingleR-HPCA (%)")
}
```

```{r confusion-celltypist, fig.width=12, fig.height=8}
if (all(c("celltypist_label", "singler_hpca") %in% colnames(seu@meta.data))) {
  ct <- table(CellTypist = seu$celltypist_label,
              HPCA       = seu$singler_hpca)
  ct_prop <- prop.table(ct, margin = 1) * 100
  pheatmap(ct_prop, fontsize = 6, display_numbers = TRUE,
           number_format = "%.0f", cluster_rows = FALSE, cluster_cols = FALSE,
           color = colorRampPalette(c("white", "steelblue"))(100),
           main = "CellTypist vs SingleR-HPCA (%)")
}
```

## 6. Composition by sample

```{r by-sample}
if ("singler_hpca" %in% colnames(seu@meta.data)) {
  df <- seu@meta.data %>%
    group_by(orig.ident, singler_hpca) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(orig.ident) %>% mutate(pct = n / sum(n) * 100)

  ggplot(df, aes(x = orig.ident, y = pct, fill = singler_hpca)) +
    geom_col() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggtitle("Cell type composition per sample (SingleR-HPCA)") +
    ylab("% cells") + xlab(NULL)
}
```

## 7. Notes / decisions

- High-confidence cell types (3 methods agree):
- Disagreements that need manual inspection:
- Clusters with low CellTypist confidence:
- Batch effect signs across samples:
- Action items before integration:

---

*Render: `quarto render 00_annotation_template.qmd -P dataset_id:GSExxxxxx --output annotation_exploration_GSExxxxxx.html`*
