---
title: "Pre-Annotation Exploration"
author: "Juliana Pinto"
format: html
editor: visual
params:
  dataset_id: "GSExxxxxx"   # ← EDIT THIS LINE before running in RStudio
---

<!--
================================================================================
HOW TO USE THIS NOTEBOOK

 1. Copy this template to a dataset-specific file:
      cp 00_annotation_template.qmd annotation_exploration_<DATASET>.qmd
    (the copy is gitignored by the **/*_GSE*.qmd rule).

 2. Open the copy in RStudio — ideally by double-clicking
    deconv_gastric_cancer.Rproj first, so the working directory is the
    repo root and here::here() resolves correctly.

 3. Edit the YAML header above:
      params:
        dataset_id: "GSE246662"   # your dataset

 4. Run cell-by-cell (Ctrl/Cmd+Enter) or render the whole notebook
    with the green "Render" button in RStudio.

 Alternative: render from the terminal without editing the YAML:
      quarto render 00_annotation_template.qmd \
        -P dataset_id:GSE246662 \
        --output annotation_exploration_GSE246662.html

 This notebook expects that scripts 09a + 09b + 09c have already been run
 for the chosen dataset. It reads annotated_seurat_path and uses
 annotation_resolution from the config. Compares SingleR-HPCA,
 SingleR-Blueprint and CellTypist side by side.
================================================================================
-->

# Pre-Annotation: `r params$dataset_id`

Compares the three pre-annotation methods (SingleR HPCA, SingleR Blueprint, CellTypist) on the output of scripts `09a` + `09b` + `09c`.

## Setup

```{r libs, message=FALSE, warning=FALSE}
library(Seurat)
library(ggplot2)
library(patchwork)
library(pheatmap)
library(dplyr)

theme_set(theme_classic())
```

```{r load-config}
# Loads config_<DATASET>.R using params$dataset_id. Paths resolve via here::here(),
# which requires this notebook to be run from the repo root (RStudio opens the
# project at the root automatically when you launch via the .Rproj file).
config_path <- file.path(
  here::here("chapter_2_sc_deconv_benchmarking", "01_sc_matrix",
             "scripts", "01_sc_pre_proc", "configs"),
  paste0("config_", params$dataset_id, ".R")
)
source(config_path)
```

```{r helper-function}
# Helper: map Ensembl IDs to gene symbols
ens_to_label <- function(ensembl_ids, seu) {
  if ("gene_symbol" %in% colnames(seu[["RNA"]]@meta.data)) {
    gene_meta <- seu[["RNA"]]@meta.data
    symbol_lookup <- setNames(gene_meta$gene_symbol, rownames(seu))
    labels <- symbol_lookup[ensembl_ids]
    labels[is.na(labels) | labels == ""] <- ensembl_ids[is.na(labels) | labels == ""]
    return(unname(labels))
  }
  return(ensembl_ids)
}
```

## 1. Load annotated object

```{r load-object}
seu <- readRDS(annotated_seurat_path)

cat("Dataset:   ", params$dataset_id, "\n")
cat("Dimensions:", nrow(seu), "genes x", ncol(seu), "cells\n")
cat("Samples:   ", length(unique(seu$orig.ident)), "\n")

# Check which annotations are available
annot_cols <- grep("singler|celltypist", colnames(seu@meta.data), value = TRUE)
cat("Annotation columns:", paste(annot_cols, collapse = ", "), "\n")
```

## 2. UMAP - All annotations

```{r umap-all, fig.width=20, fig.height=16}
annotation_cluster_col <- paste0("RNA_snn_res.", annotation_resolution)

p1 <- DimPlot(seu, group.by = annotation_cluster_col, label = TRUE) +
  ggtitle(paste("Clusters (res", annotation_resolution, ")")) + NoLegend()

plots <- list(p1)

if ("singler_hpca" %in% colnames(seu@meta.data)) {
  plots[["hpca"]] <- DimPlot(seu, group.by = "singler_hpca", label = TRUE, repel = TRUE) +
    ggtitle("SingleR - HPCA") + NoLegend()
}

if ("singler_blueprint" %in% colnames(seu@meta.data)) {
  plots[["bp"]] <- DimPlot(seu, group.by = "singler_blueprint", label = TRUE, repel = TRUE) +
    ggtitle("SingleR - Blueprint") + NoLegend()
}

if ("celltypist_label" %in% colnames(seu@meta.data)) {
  plots[["ct"]] <- DimPlot(seu, group.by = "celltypist_label", label = TRUE, repel = TRUE) +
    ggtitle("CellTypist") + NoLegend()
}

wrap_plots(plots, ncol = 2)
```

## 3. Annotation counts

```{r annotation-counts}
if ("singler_hpca" %in% colnames(seu@meta.data)) {
  cat("--- SingleR HPCA ---\n")
  print(sort(table(seu$singler_hpca), decreasing = TRUE))
  cat("\n")
}

if ("singler_blueprint" %in% colnames(seu@meta.data)) {
  cat("--- SingleR Blueprint ---\n")
  print(sort(table(seu$singler_blueprint), decreasing = TRUE))
  cat("\n")
}

if ("celltypist_label" %in% colnames(seu@meta.data)) {
  cat("--- CellTypist ---\n")
  print(sort(table(seu$celltypist_label), decreasing = TRUE))
}
```

## 4. CellTypist confidence scores

```{r celltypist-confidence-umap, fig.width=8, fig.height=7}
if ("celltypist_conf_score" %in% colnames(seu@meta.data)) {
  FeaturePlot(seu, features = "celltypist_conf_score", order = TRUE) +
    scale_color_gradient(low = "red", high = "steelblue", name = "Score") +
    ggtitle("CellTypist confidence on UMAP")
}
```

```{r celltypist-confidence-hist, fig.width=10, fig.height=4}
if ("celltypist_conf_score" %in% colnames(seu@meta.data)) {
  ggplot(seu@meta.data, aes(x = celltypist_conf_score)) +
    geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
    geom_vline(xintercept = 0.5, linetype = "dashed", color = "red") +
    ggtitle("CellTypist confidence distribution") +
    xlab("Confidence score") + ylab("Cells")
}
```

```{r low-confidence}
if ("celltypist_conf_score" %in% colnames(seu@meta.data)) {
  low_conf <- sum(seu$celltypist_conf_score < 0.5, na.rm = TRUE)
  cat("Cells with confidence < 0.5:", low_conf,
      "(", round(100 * low_conf / ncol(seu), 1), "%)\n")

  # Which labels have lowest confidence?
  conf_by_label <- seu@meta.data %>%
    group_by(celltypist_label) %>%
    summarise(
      n         = n(),
      mean_conf = round(mean(celltypist_conf_score, na.rm = TRUE), 3),
      pct_low   = round(sum(celltypist_conf_score < 0.5, na.rm = TRUE) / n() * 100, 1),
      .groups   = "drop"
    ) %>%
    arrange(mean_conf)

  knitr::kable(conf_by_label, caption = "CellTypist confidence by label")
}
```

## 5. Cluster-level consensus

```{r cluster-consensus}
annotation_cluster_col <- paste0("RNA_snn_res.", annotation_resolution)

consensus <- seu@meta.data %>%
  group_by(across(all_of(annotation_cluster_col))) %>%
  summarise(
    n_cells        = n(),
    hpca           = if ("singler_hpca" %in% colnames(seu@meta.data))
                       names(sort(table(singler_hpca), decreasing = TRUE))[1] else NA,
    hpca_pct       = if ("singler_hpca" %in% colnames(seu@meta.data))
                       round(max(table(singler_hpca)) / n() * 100, 1) else NA,
    blueprint      = if ("singler_blueprint" %in% colnames(seu@meta.data))
                       names(sort(table(singler_blueprint), decreasing = TRUE))[1] else NA,
    blueprint_pct  = if ("singler_blueprint" %in% colnames(seu@meta.data))
                       round(max(table(singler_blueprint)) / n() * 100, 1) else NA,
    celltypist     = if ("celltypist_label" %in% colnames(seu@meta.data))
                       names(sort(table(celltypist_label), decreasing = TRUE))[1] else NA,
    celltypist_pct = if ("celltypist_label" %in% colnames(seu@meta.data))
                       round(max(table(celltypist_label)) / n() * 100, 1) else NA,
    .groups = "drop"
  )

knitr::kable(consensus, caption = "Dominant annotation per cluster (% agreement)")
```

## 6. Confusion matrices

### SingleR HPCA vs Clusters

```{r confusion-hpca, fig.width=14, fig.height=8}
if ("singler_hpca" %in% colnames(seu@meta.data)) {
  ct <- table(
    Cluster = seu@meta.data[[annotation_cluster_col]],
    HPCA    = seu$singler_hpca
  )
  ct_prop <- prop.table(ct, margin = 1) * 100

  pheatmap(
    ct_prop,
    fontsize        = 7,
    display_numbers = TRUE,
    number_format   = "%.0f",
    cluster_rows    = FALSE,
    cluster_cols    = FALSE,
    color           = colorRampPalette(c("white", "steelblue"))(100),
    main            = "Cluster vs SingleR-HPCA (%)"
  )
}
```

### SingleR Blueprint vs Clusters

```{r confusion-blueprint, fig.width=14, fig.height=8}
if ("singler_blueprint" %in% colnames(seu@meta.data)) {
  ct <- table(
    Cluster   = seu@meta.data[[annotation_cluster_col]],
    Blueprint = seu$singler_blueprint
  )
  ct_prop <- prop.table(ct, margin = 1) * 100

  pheatmap(
    ct_prop,
    fontsize        = 7,
    display_numbers = TRUE,
    number_format   = "%.0f",
    cluster_rows    = FALSE,
    cluster_cols    = FALSE,
    color           = colorRampPalette(c("white", "steelblue"))(100),
    main            = "Cluster vs SingleR-Blueprint (%)"
  )
}
```

### CellTypist vs SingleR HPCA

```{r confusion-ct-hpca, fig.width=16, fig.height=10}
if (all(c("celltypist_label", "singler_hpca") %in% colnames(seu@meta.data))) {
  ct <- table(
    CellTypist = seu$celltypist_label,
    HPCA       = seu$singler_hpca
  )
  ct_prop <- prop.table(ct, margin = 1) * 100

  pheatmap(
    ct_prop,
    fontsize        = 6,
    display_numbers = TRUE,
    number_format   = "%.0f",
    cluster_rows    = FALSE,
    cluster_cols    = FALSE,
    color           = colorRampPalette(c("white", "steelblue"))(100),
    main            = "CellTypist vs SingleR-HPCA (%)"
  )
}
```

### CellTypist vs SingleR Blueprint

```{r confusion-ct-bp, fig.width=16, fig.height=10}
if (all(c("celltypist_label", "singler_blueprint") %in% colnames(seu@meta.data))) {
  ct <- table(
    CellTypist = seu$celltypist_label,
    Blueprint  = seu$singler_blueprint
  )
  ct_prop <- prop.table(ct, margin = 1) * 100

  pheatmap(
    ct_prop,
    fontsize        = 6,
    display_numbers = TRUE,
    number_format   = "%.0f",
    cluster_rows    = FALSE,
    cluster_cols    = FALSE,
    color           = colorRampPalette(c("white", "steelblue"))(100),
    main            = "CellTypist vs SingleR-Blueprint (%)"
  )
}
```

## 7. By sample

```{r annotation-by-sample, fig.width=14, fig.height=6}
if ("singler_hpca" %in% colnames(seu@meta.data)) {
  df <- seu@meta.data %>%
    group_by(orig.ident, singler_hpca) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(orig.ident) %>%
    mutate(pct = n / sum(n) * 100)

  ggplot(df, aes(x = orig.ident, y = pct, fill = singler_hpca)) +
    geom_col() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggtitle("Cell type composition per sample (SingleR-HPCA)") +
    ylab("% cells") + xlab(NULL)
}
```

## 8. Summary

```{r summary}
cat("Dataset:    ", params$dataset_id, "\n")
cat("Cells:      ", ncol(seu), "\n")
cat("Resolution: ", annotation_resolution, "\n")
cat("Annotations available:", paste(annot_cols, collapse = ", "), "\n")

if ("singler_hpca" %in% colnames(seu@meta.data))
  cat("SingleR HPCA labels:     ", length(unique(seu$singler_hpca)), "\n")
if ("singler_blueprint" %in% colnames(seu@meta.data))
  cat("SingleR Blueprint labels:", length(unique(seu$singler_blueprint)), "\n")
if ("celltypist_label" %in% colnames(seu@meta.data)) {
  cat("CellTypist labels:       ", length(unique(na.omit(seu$celltypist_label))), "\n")
  cat("CellTypist mean confidence:",
      round(mean(seu$celltypist_conf_score, na.rm = TRUE), 3), "\n")
}
```

## 9. Conclusions

Notes (fill in while inspecting the plots above):

-   **Agreement**: where do all three methods agree? Those are high-confidence cell type assignments.
-   **Disagreements**: where they disagree, which method seems more plausible given the known markers?
-   **CellTypist confidence**: cells with \< 0.5 confidence may be misclassified or represent transitional states.
-   **Batch effect**: do annotations vary by sample? This indicates batch effect still present (expected pre-integration).
-   **Epithelial cells**: CellTypist (immune-focused model) may struggle with epithelial cells — SingleR with HPCA is better for these.
-   **Reminder**: this is pre-annotation. Final annotation will be done on the integrated object.
