---
title: "Single-cell dataset exploration"
author: "Juliana Pinto"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
params:
  dataset_id: "GSExxxxxx"
  data_dir: "data/sc_reference/raw/GSExxxxxx"
  # --- Format: choose ONE ---
  # "csv"  -> CSV files (one per sample)
  # "10x"  -> 10x Genomics directories (barcodes/features/matrix)
  # "h5"   -> Single HDF5 file
  # "rds"  -> RDS file (Seurat object or matrix)
  data_format: "10x"
  # --- Only for h5/rds: filename within data_dir ---
  data_file: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

# Dataset: `r params$dataset_id`

This notebook explores the basic characteristics of a single-cell RNA-seq
dataset prior to preprocessing. The goal is to understand the data structure,
identify the types of genes present, and check the availability of metadata.

## Loading data

```{r load-libraries}
library(Seurat)
library(Matrix)

project_root <- here::here()
data_path <- file.path(project_root, params$data_dir)
```

```{r load-data}
# -------------------------------------------------------------------
# Load data according to format
# Output: counts_list (named list of count matrices, one per sample)
#         is_seurat    (TRUE if the source was a Seurat object)
#         seurat_obj   (the original object, only when format = "rds")
# -------------------------------------------------------------------
is_seurat <- FALSE
seurat_obj <- NULL

if (params$data_format == "csv") {

  csv_files <- list.files(data_path, pattern = "\\.csv(\\.gz)?$",
                          full.names = TRUE)
  cat("CSV files found:", length(csv_files), "\n")
  cat(paste("-", basename(csv_files)), sep = "\n")

  counts_list <- list()
  for (f in csv_files) {
    sample_name <- sub("\\.csv(\\.gz)?$", "", basename(f))
    cat("Reading:", sample_name, "...\n")
    counts_list[[sample_name]] <- read.csv(f, row.names = 1,
                                           check.names = FALSE)
  }

} else if (params$data_format == "10x") {

  sample_dirs <- list.dirs(data_path, recursive = FALSE, full.names = TRUE)
  # Keep only directories that contain 10x files
  sample_dirs <- sample_dirs[sapply(sample_dirs, function(d) {
    file.exists(file.path(d, "matrix.mtx.gz"))
  })]
  cat("10x sample directories found:", length(sample_dirs), "\n")
  cat(paste("-", basename(sample_dirs)), sep = "\n")

  counts_list <- list()
  failed_samples <- character()
  for (d in sample_dirs) {
    sample_name <- basename(d)
    cat("Reading:", sample_name, "...")
    result <- tryCatch(Read10X(data.dir = d),
                       error = function(e) {
                         cat(" FAILED -", conditionMessage(e), "\n")
                         return(NULL)
                       })
    if (!is.null(result)) {
      counts_list[[sample_name]] <- result
      cat(" OK\n")
    } else {
      failed_samples <- c(failed_samples, sample_name)
    }
  }
  if (length(failed_samples) > 0) {
    cat("\nFailed samples:\n")
    cat(paste("-", failed_samples), sep = "\n")
  }

} else if (params$data_format == "h5") {

  h5_path <- file.path(data_path, params$data_file)
  cat("H5 file:", basename(h5_path), "\n")
  cat("File size:", round(file.size(h5_path) / 1e6, 1), "MB\n")

  counts <- Read10X_h5(filename = h5_path)
  if (is.list(counts)) {
    cat("Multiple modalities detected:", paste(names(counts), collapse = ", "), "\n")
    cat("Using 'Gene Expression' matrix.\n")
    counts <- counts[["Gene Expression"]]
  }
  counts_list <- list(all_cells = counts)
  rm(counts)

} else if (params$data_format == "rds") {

  rds_path <- file.path(data_path, params$data_file)
  cat("RDS file:", basename(rds_path), "\n")
  cat("File size:", round(file.size(rds_path) / 1e9, 2), "GB\n")

  obj <- readRDS(rds_path)
  if (inherits(obj, "Seurat")) {
    obj <- UpdateSeuratObject(obj)
    is_seurat <- TRUE
    seurat_obj <- obj
    cat("Seurat object loaded.\n")
    print(obj)
    counts <- tryCatch(
      GetAssayData(obj, assay = "RNA", slot = "counts"),
      error = function(e) GetAssayData(obj, slot = "counts")
    )
  } else {
    counts <- as(obj, "dgCMatrix")
    rm(obj)
  }
  counts_list <- list(all_cells = counts)
  rm(counts)

} else {
  stop("Unknown data_format: ", params$data_format,
       ". Use 'csv', '10x', 'h5', or 'rds'.")
}

cat("\nSamples loaded:", length(counts_list), "\n")
cat("Total cells:", sum(sapply(counts_list, ncol)), "\n")
```

---

## Per-sample exploration

The sections below apply each check across all samples in the dataset.

### Matrix dimensions (`dim`)

We check the number of genes (rows) and cells (columns) for each sample.
The expected format is **genes x cells**.

```{r dim-counts}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  cat("\n===", name, "===\n")
  cat("dim(counts):", dim(counts), "\n")
  cat("  -> Genes:", nrow(counts), " | Cells:", ncol(counts), "\n")
}
```

### Gene identifiers (`head(rownames)`)

We inspect the first gene names to identify the type of annotation used
(gene symbols, Ensembl IDs, etc.).

```{r head-rownames}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  cat("\n===", name, "===\n")
  cat("head(rownames(counts)):\n")
  print(head(rownames(counts)))
}
```

### First 20 genes (`head(rownames, 20)`)

An expanded listing of the first 20 genes for a more detailed inspection
of the naming convention.

```{r head-rownames-20}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  cat("\n===", name, "===\n")
  cat("head(rownames(counts), 20):\n")
  print(head(rownames(counts), 20))
}
```

```{r gene-id-type}
first_genes <- head(rownames(counts_list[[1]]), 10)

gene_id_type <- if (any(grepl("^ENSG", first_genes))) {
  "Ensembl"
} else if (any(grepl("^[A-Z][A-Z0-9]+$", first_genes))) {
  "Gene Symbol"
} else {
  "Unknown"
}

cat("Detected gene identifier type:", gene_id_type, "\n")
```

### Cell barcodes (`head(colnames)`)

We check the format of cell barcodes (matrix columns). Typical 10x Genomics
barcodes follow the pattern `ACGT...ACGT-1`.

```{r head-colnames}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  cat("\n===", name, "===\n")
  cat("head(colnames(counts)):\n")
  print(head(colnames(counts)))
}
```

### Sequencing depth (`summary(colSums)`)

The library size (total UMI count per cell) indicates sequencing depth.
Very low values may indicate low-quality cells (empty droplets).

```{r summary-colsums}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  cat("\n===", name, "===\n")
  cat("summary(colSums(counts)):\n")
  print(summary(colSums(counts)))
}
```

### Genes detected per cell (`summary(colSums(counts > 0))`)

The number of genes with at least 1 UMI per cell reflects transcriptional
complexity. Cells with very few detected genes may be of low quality.

```{r summary-genes-detected}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  cat("\n===", name, "===\n")
  cat("summary(colSums(counts > 0)):\n")
  print(summary(colSums(counts > 0)))
}
```

### Sequencing depth histogram (`hist(log10(colSums + 1))`)

The log10-scaled distribution allows visualizing the separation between
empty droplets and real cells. A bimodal or unimodal distribution with a
left tail is expected.

```{r hist-colsums, fig.height=4, results='asis'}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  hist(
    log10(colSums(counts) + 1),
    main = paste("Library size -", name),
    xlab = "log10(Total UMI + 1)",
    col = "steelblue",
    border = "white",
    breaks = 50
  )
  cat("\n\n")
}
```

### Mitochondrial genes (`grepl("^MT-")`)

Mitochondrial genes (`MT-` prefix) are classic indicators of cellular stress.
A high fraction of mitochondrial reads (>20%) typically indicates cells
undergoing apoptosis or damaged during tissue dissociation.

```{r mito-genes}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  mt_count <- sum(grepl("^MT-", rownames(counts)))
  cat("\n===", name, "===\n")
  cat("sum(grepl('^MT-', rownames(counts))):", mt_count, "\n")
  if (mt_count > 0) {
    cat("MT genes found:\n")
    print(grep("^MT-", rownames(counts), value = TRUE))
  }
}
```

### Ribosomal genes (`grepl("^RPL|^RPS")`)

Ribosomal genes (prefixes `RPL` — large ribosomal subunit proteins — and
`RPS` — small subunit) represent a significant fraction of the transcriptome.
High fractions can mask relevant biological variability and are often filtered
out or regressed.

```{r ribo-genes}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  ribo_count <- sum(grepl("^RPL|^RPS", rownames(counts)))
  cat("\n===", name, "===\n")
  cat("sum(grepl('^RPL|^RPS', rownames(counts))):", ribo_count, "\n")
}
```

### Hemoglobin genes (`grepl("^HB")`)

Hemoglobin genes (e.g., `HBA1`, `HBA2`, `HBB`) indicate red blood cell
contamination. Their excessive presence suggests issues with tissue
dissociation or erythrocyte lysis.

```{r hb-genes}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  hb_count <- sum(grepl("^HB[ABDEG]", rownames(counts)))
  cat("\n===", name, "===\n")
  cat("sum(grepl('^HB', rownames(counts))):", hb_count, "\n")
  if (hb_count > 0) {
    cat("HB genes found:\n")
    print(grep("^HB[ABDEG]", rownames(counts), value = TRUE))
  }
}
```

### Duplicated genes (`sum(duplicated(rownames))`)

Duplicated gene names cause issues when creating Seurat/SingleCellExperiment
objects and in downstream analyses. If present, they should be handled
(e.g., `make.unique()`).

```{r duplicated-genes}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  dup_count <- sum(duplicated(rownames(counts)))
  cat("\n===", name, "===\n")
  cat("sum(duplicated(rownames(counts))):", dup_count, "\n")
  if (dup_count > 0) {
    dups <- rownames(counts)[duplicated(rownames(counts))]
    cat("First duplicated genes:\n")
    print(head(dups, 10))
  }
}
```

### Metadata check

We check whether metadata is available — either embedded in a Seurat object
or as separate files in the dataset directory.

```{r metadata-check}
meta <- NULL

if (is_seurat && !is.null(seurat_obj)) {
  meta <- seurat_obj@meta.data
  cat("Metadata from Seurat object.\n")
  cat("Dimensions:", dim(meta), "\n")
  cat("  -> Cells:", nrow(meta), " | Fields:", ncol(meta), "\n\n")
  cat("Available metadata columns:\n")
  cat(paste(" ", colnames(meta)), sep = "\n")
} else {
  meta_files <- list.files(
    data_path,
    pattern = "meta|annotation|celltype|cluster|barcode",
    ignore.case = TRUE,
    full.names = TRUE,
    recursive = TRUE
  )
  # Exclude standard 10x barcodes files
  meta_files <- meta_files[!grepl("barcodes\\.tsv\\.gz$", meta_files)]

  cat("Metadata files found:", length(meta_files), "\n")
  if (length(meta_files) > 0) {
    cat(paste("-", basename(meta_files)), sep = "\n")
  }
}
```

```{r metadata-fields}
expected_fields <- c("celltype", "cell_type", "cell.type",
                     "patient", "donor", "subject",
                     "tissue", "organ",
                     "condition", "disease", "treatment",
                     "sample", "cluster", "annotation",
                     "orig.ident", "seurat_clusters")

if (!is.null(meta)) {
  # Seurat embedded metadata
  matches <- colnames(meta)[tolower(colnames(meta)) %in% expected_fields]
  cat("Fields of interest found:",
      ifelse(length(matches) > 0, paste(matches, collapse = ", "), "none"),
      "\n\n")
  for (field in matches) {
    cat("\n===", field, "===\n")
    vals <- unique(meta[[field]])
    cat("Unique values:", length(vals), "\n")
    if (length(vals) <= 30) {
      print(sort(vals))
    } else {
      cat("First 30 values:\n")
      print(head(sort(vals), 30))
    }
  }
} else if (exists("meta_files") && length(meta_files) > 0) {
  # Separate metadata files
  for (mf in meta_files) {
    cat("\n===", basename(mf), "===\n")
    mdata <- tryCatch(read.csv(mf, nrows = 5), error = function(e) NULL)
    if (!is.null(mdata)) {
      cat("Available columns:\n")
      cat(paste(" ", colnames(mdata)), sep = "\n")
      matches <- intersect(tolower(colnames(mdata)), expected_fields)
      cat("\nFields of interest found:",
          ifelse(length(matches) > 0, paste(matches, collapse = ", "), "none"),
          "\n")
    }
  }
} else {
  cat("No metadata available.\n")
  cat("  - celltype:  NOT AVAILABLE\n")
  cat("  - patient:   NOT AVAILABLE\n")
  cat("  - tissue:    NOT AVAILABLE\n")
  cat("  - condition: NOT AVAILABLE\n")
}
```

### Per-sample cell counts

If sample-level identifiers exist in the metadata, we report the number of
cells per sample.

```{r per-sample-counts}
sample_counts <- NULL

if (!is.null(meta)) {
  sample_col <- NULL
  for (col in c("orig.ident", "sample", "patient", "donor", "subject")) {
    if (col %in% colnames(meta)) {
      sample_col <- col
      break
    }
  }
  if (!is.null(sample_col)) {
    cat("Using column '", sample_col, "' as sample identifier:\n\n", sep = "")
    sample_counts <- sort(table(meta[[sample_col]]), decreasing = TRUE)
    print(data.frame(
      Sample = names(sample_counts),
      Cells = as.integer(sample_counts),
      row.names = NULL
    ))
    cat("\nTotal samples:", length(sample_counts), "\n")
  }
} else if (length(counts_list) > 1) {
  cat("Samples from file structure:\n\n")
  sample_counts <- sapply(counts_list, ncol)
  print(data.frame(
    Sample = names(sample_counts),
    Cells = as.integer(sample_counts),
    row.names = NULL
  ))
  cat("\nTotal samples:", length(sample_counts), "\n")
} else {
  cat("Single matrix loaded; no per-sample breakdown available.\n")
}
```

---

## Overall dataset summary

Summary table with the main characteristics identified in this exploration.

```{r cleanup, include=FALSE}
# Save summary variables before freeing large objects
first_counts <- counts_list[[1]]
mt_present   <- any(sapply(counts_list, function(x) sum(grepl("^MT-", rownames(x)))) > 0)
ribo_present <- any(sapply(counts_list, function(x) sum(grepl("^RPL|^RPS", rownames(x)))) > 0)
hb_present   <- any(sapply(counts_list, function(x) sum(grepl("^HB[ABDEG]", rownames(x)))) > 0)
dup_present  <- any(sapply(counts_list, function(x) sum(duplicated(rownames(x)))) > 0)
n_genes      <- nrow(first_counts)
n_cells      <- sum(sapply(counts_list, ncol))
n_samples    <- if (!is.null(sample_counts)) length(sample_counts) else length(counts_list)

meta_available <- if (!is.null(meta)) {
  key_fields <- colnames(meta)[tolower(colnames(meta)) %in%
    c("celltype", "cell_type", "cell.type", "annotation", "cluster", "seurat_clusters")]
  ifelse(length(key_fields) > 0,
         paste("Available (", paste(key_fields, collapse = ", "), ")", sep = ""),
         "Partial (no cell type annotation)")
} else if (exists("meta_files") && length(meta_files) > 0) {
  "Available (separate files)"
} else {
  "Not available"
}

# Free memory (important for large datasets like RDS)
rm(counts_list, first_counts)
if (exists("seurat_obj")) rm(seurat_obj)
if (exists("meta")) rm(meta)
gc()
```

```{r summary-table}
summary_info <- data.frame(
  Field = c(
    "Dataset",
    "Data format",
    "Samples (libraries)",
    "Total genes",
    "Total cells",
    "Gene ID type",
    "Mitochondrial genes",
    "Ribosomal genes",
    "Hemoglobin genes",
    "Duplicated genes",
    "Metadata available"
  ),
  Value = c(
    params$dataset_id,
    params$data_format,
    as.character(n_samples),
    n_genes,
    n_cells,
    gene_id_type,
    ifelse(mt_present, "Present", "Absent"),
    ifelse(ribo_present, "Present", "Absent"),
    ifelse(hb_present, "Present", "Absent"),
    ifelse(dup_present, "Present", "Absent"),
    meta_available
  )
)

knitr::kable(summary_info, caption = paste("Dataset summary -", params$dataset_id))
```

---

*Single-cell RNA-seq dataset initial exploration notebook.*
