---
title: "Single-cell dataset exploration"
author: "Juliana Pinto"
date: 03/12/2026
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
params:
  dataset_id: "GSE163558"
  data_dir: "data/sc_reference/raw/GSE163558"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

# Dataset: `r params$dataset_id`

This notebook explores the basic characteristics of a single-cell RNA-seq dataset prior to preprocessing. The goal is to understand the data structure, identify the types of genes present, and check the availability of metadata.

This dataset is in **10x Genomics** format (barcodes.tsv.gz, features.tsv.gz, matrix.mtx.gz), so we use `Seurat::Read10X()` to load sparse count matrices.

## Loading files

We identify all sample subdirectories containing 10x Genomics files and load each one as a sparse count matrix.

```{r load-files}
library(Seurat)
library(Matrix)

project_root <- here::here()
data_path <- file.path(project_root, params$data_dir)

# Each sample is in its own subdirectory with barcodes/features/matrix files
# Filter only directories containing 10x files (excludes FIXED, processed, etc.)
sample_dirs <- list.dirs(data_path, recursive = FALSE, full.names = TRUE)
sample_dirs <- sample_dirs[sapply(sample_dirs, function(d) {
  file.exists(file.path(d, "matrix.mtx.gz"))
})]

cat("Number of samples found:", length(sample_dirs), "\n")
cat("Samples:\n")
cat(paste("-", basename(sample_dirs)), sep = "\n")
```

```{r read-all}
counts_list <- list()

for (d in sample_dirs) {
  sample_name <- basename(d)
  cat("Reading:", sample_name, "...\n")
  counts_list[[sample_name]] <- Read10X(data.dir = d)
}
```

---

## Per-sample exploration

The sections below apply each check individually across all samples in the dataset.

### Matrix dimensions (`dim`)

We check the number of genes (rows) and cells (columns) for each sample. The expected format is **genes × cells**.

```{r dim-counts}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  cat("\n===", name, "===\n")
  cat("dim(counts):", dim(counts), "\n")
  cat("  -> Genes:", nrow(counts), " | Cells:", ncol(counts), "\n")
}
```

### Gene identifiers (`head(rownames)`)

We inspect the first gene names to identify the type of annotation used (gene symbols, Ensembl IDs, etc.).

```{r head-rownames}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  cat("\n===", name, "===\n")
  cat("head(rownames(counts)):\n")
  print(head(rownames(counts)))
}
```

### First 20 genes (`head(rownames, 20)`)

An expanded listing of the first 20 genes for a more detailed inspection of the naming convention.

```{r head-rownames-20}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  cat("\n===", name, "===\n")
  cat("head(rownames(counts), 20):\n")
  print(head(rownames(counts), 20))
}
```

```{r gene-id-type}
# Automatic detection of gene identifier type
first_genes <- head(rownames(counts_list[[1]]), 10)

gene_id_type <- if (any(grepl("^ENSG", first_genes))) {
  "Ensembl"
} else if (any(grepl("^[A-Z][A-Z0-9]+$", first_genes))) {
  "Gene Symbol"
} else {
  "Unknown"
}

cat("Detected gene identifier type:", gene_id_type, "\n")
```

### Cell barcodes (`head(colnames)`)

We check the format of cell barcodes (matrix columns). Typical 10x Genomics barcodes follow the pattern `ACGT...ACGT-1`.

```{r head-colnames}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  cat("\n===", name, "===\n")
  cat("head(colnames(counts)):\n")
  print(head(colnames(counts)))
}
```

### Sequencing depth (`summary(colSums)`)

The library size (total UMI count per cell) indicates sequencing depth. Very low values may indicate low-quality cells (empty droplets).

```{r summary-colsums}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  cat("\n===", name, "===\n")
  cat("summary(colSums(counts)):\n")
  print(summary(colSums(counts)))
}
```

### Genes detected per cell (`summary(colSums(counts > 0))`)

The number of genes with at least 1 UMI per cell reflects transcriptional complexity. Cells with very few detected genes may be of low quality.

```{r summary-genes-detected}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  cat("\n===", name, "===\n")
  cat("summary(colSums(counts > 0)):\n")
  print(summary(colSums(counts > 0)))
}
```

### Sequencing depth histogram (`hist(log10(colSums + 1))`)

The log10-scaled distribution allows visualizing the separation between empty droplets and real cells. A bimodal or unimodal distribution with a left tail is expected.

```{r hist-colsums, fig.height=4, results='asis'}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  hist(
    log10(colSums(counts) + 1),
    main = paste("Library size -", name),
    xlab = "log10(Total UMI + 1)",
    col = "steelblue",
    border = "white",
    breaks = 50
  )
  cat("\n\n")
}
```

### Mitochondrial genes (`grepl("^MT-")`)

Mitochondrial genes (`MT-` prefix) are classic indicators of cellular stress. A high fraction of mitochondrial reads (>20%) typically indicates cells undergoing apoptosis or damaged during tissue dissociation.

```{r mito-genes}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  mt_count <- sum(grepl("^MT-", rownames(counts)))
  cat("\n===", name, "===\n")
  cat("sum(grepl('^MT-', rownames(counts))):", mt_count, "\n")
  if (mt_count > 0) {
    cat("MT genes found:\n")
    print(grep("^MT-", rownames(counts), value = TRUE))
  }
}
```

### Ribosomal genes (`grepl("^RPL|^RPS")`)

Ribosomal genes (prefixes `RPL` — large ribosomal subunit proteins — and `RPS` — small subunit) represent a significant fraction of the transcriptome. High fractions can mask relevant biological variability and are often filtered out or regressed.

```{r ribo-genes}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  ribo_count <- sum(grepl("^RPL|^RPS", rownames(counts)))
  cat("\n===", name, "===\n")
  cat("sum(grepl('^RPL|^RPS', rownames(counts))):", ribo_count, "\n")
}
```

### Hemoglobin genes (`grepl("^HB")`)

Hemoglobin genes (e.g., `HBA1`, `HBA2`, `HBB`) indicate red blood cell contamination. Their excessive presence suggests issues with tissue dissociation or erythrocyte lysis.

```{r hb-genes}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  hb_count <- sum(grepl("^HB[ABDEG]", rownames(counts)))
  cat("\n===", name, "===\n")
  cat("sum(grepl('^HB', rownames(counts))):", hb_count, "\n")
  if (hb_count > 0) {
    cat("HB genes found:\n")
    print(grep("^HB[ABDEG]", rownames(counts), value = TRUE))
  }
}
```

### Duplicated genes (`sum(duplicated(rownames))`)

Duplicated gene names cause issues when creating Seurat/SingleCellExperiment objects and in downstream analyses. If present, they should be handled (e.g., `make.unique()`).

```{r duplicated-genes}
for (name in names(counts_list)) {
  counts <- counts_list[[name]]
  dup_count <- sum(duplicated(rownames(counts)))
  cat("\n===", name, "===\n")
  cat("sum(duplicated(rownames(counts))):", dup_count, "\n")
  if (dup_count > 0) {
    dups <- rownames(counts)[duplicated(rownames(counts))]
    cat("First duplicated genes:\n")
    print(head(dups, 10))
  }
}
```

### Metadata check

We check whether metadata files exist in the dataset directory and whether they contain relevant fields such as **cell type**, **patient**, **tissue**, and **condition**.

```{r metadata-check}
meta_files <- list.files(
  data_path,
  pattern = "meta|annotation|celltype|cluster|barcode",
  ignore.case = TRUE,
  full.names = TRUE,
  recursive = TRUE
)

# Exclude the standard 10x barcodes.tsv.gz files
meta_files <- meta_files[!grepl("barcodes\\.tsv\\.gz$", meta_files)]

cat("Metadata files found:", length(meta_files), "\n")
if (length(meta_files) > 0) {
  cat(paste("-", basename(meta_files)), sep = "\n")
}
```

```{r metadata-fields}
expected_fields <- c("celltype", "cell_type", "cell.type",
                     "patient", "donor", "subject",
                     "tissue", "organ",
                     "condition", "disease", "treatment",
                     "sample", "cluster", "annotation")

if (length(meta_files) > 0) {
  for (mf in meta_files) {
    cat("\n===", basename(mf), "===\n")
    meta <- tryCatch(
      read.csv(mf, nrows = 5),
      error = function(e) NULL
    )
    if (!is.null(meta)) {
      cat("Available columns:\n")
      cat(paste(" ", colnames(meta)), sep = "\n")
      matches <- intersect(tolower(colnames(meta)), expected_fields)
      cat("\nFields of interest found:",
          ifelse(length(matches) > 0, paste(matches, collapse = ", "), "none"),
          "\n")
    }
  }
} else {
  cat("No separate metadata files found.\n")
  cat("\nMetadata availability:\n")
  cat("  - celltype:  NOT AVAILABLE\n")
  cat("  - patient:   NOT AVAILABLE\n")
  cat("  - tissue:    NOT AVAILABLE\n")
  cat("  - condition: NOT AVAILABLE\n")
}
```

---

## Overall dataset summary

Summary table with the main characteristics identified in this exploration.

```{r summary-table}
first_counts <- counts_list[[1]]

mt_present   <- any(sapply(counts_list, function(x) sum(grepl("^MT-", rownames(x)))) > 0)
ribo_present <- any(sapply(counts_list, function(x) sum(grepl("^RPL|^RPS", rownames(x)))) > 0)
hb_present   <- any(sapply(counts_list, function(x) sum(grepl("^HB[ABDEG]", rownames(x)))) > 0)
dup_present  <- any(sapply(counts_list, function(x) sum(duplicated(rownames(x)))) > 0)

summary_info <- data.frame(
  Field = c(
    "Dataset",
    "Samples (libraries)",
    "Total genes",
    "Total cells",
    "Gene ID type",
    "Mitochondrial genes",
    "Ribosomal genes",
    "Hemoglobin genes",
    "Duplicated genes",
    "Metadata available"
  ),
  Value = c(
    params$dataset_id,
    length(counts_list),
    nrow(first_counts),
    sum(sapply(counts_list, ncol)),
    gene_id_type,
    ifelse(mt_present, "Present", "Absent"),
    ifelse(ribo_present, "Present", "Absent"),
    ifelse(hb_present, "Present", "Absent"),
    ifelse(dup_present, "Present", "Absent"),
    ifelse(length(meta_files) > 0, "Available", "Not available")
  )
)

knitr::kable(summary_info, caption = paste("Dataset summary -", params$dataset_id))
```

---

*Single-cell RNA-seq dataset initial exploration notebook.*
