---
title: "Single-cell dataset exploration"
author: "Juliana Pinto"
date: 03/13/2026
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
params:
  dataset_id: "GSE201347"
  data_dir: "data/sc_reference/raw/GSE201347"
  data_file: "GSE201347.rds"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

# Dataset: `r params$dataset_id`

This notebook explores the basic characteristics of a single-cell RNA-seq dataset prior to preprocessing. The goal is to understand the data structure, identify the types of genes present, and check the availability of metadata.

This dataset is provided as an **RDS file** containing a pre-built Seurat object. The original filename (`GC.integrated.no.NAC.rds`) suggests it is an already integrated Seurat object from gastric cancer samples (excluding neoadjuvant chemotherapy patients).

## Loading the RDS file

We load the RDS file and inspect the object class. Since the file is large (~18 GB), this step may take several minutes.

```{r load-file}
# Increase memory limit for this large dataset (~18 GB)
Sys.setenv("R_MAX_VSIZE" = "64Gb")

library(Seurat)
library(Matrix)

project_root <- here::here()
data_path <- file.path(project_root, params$data_dir)
file_path <- file.path(data_path, params$data_file)

cat("Loading:", basename(file_path), "\n")
cat("File size:", round(file.size(file_path) / 1e9, 2), "GB\n")

obj <- readRDS(file_path)

cat("Object class:", class(obj), "\n")
```

## Object structure

We inspect the overall structure of the loaded object to understand what data and metadata are available.

```{r object-structure}
if (inherits(obj, "Seurat")) {
  cat("This is a Seurat object.\n\n")
  print(obj)
} else {
  cat("Object class:", class(obj), "\n")
  cat("Object structure:\n")
  str(obj, max.level = 2)
}
```

## Extracting the count matrix

We extract the raw count matrix from the object. For Seurat objects we look for the RNA assay `counts` slot; otherwise we attempt to use the object directly as a matrix.

```{r extract-counts}
if (inherits(obj, "Seurat")) {
  # Check available assays
  cat("Available assays:", paste(Assays(obj), collapse = ", "), "\n")

  # Try to get raw counts from the RNA assay
  default_assay <- DefaultAssay(obj)
  cat("Default assay:", default_assay, "\n\n")

  counts <- tryCatch(
    GetAssayData(obj, assay = "RNA", slot = "counts"),
    error = function(e) {
      cat("Could not get counts from RNA assay, trying default assay...\n")
      GetAssayData(obj, slot = "counts")
    }
  )
} else if (inherits(obj, c("dgCMatrix", "matrix", "Matrix"))) {
  counts <- obj
} else {
  cat("Unexpected object type:", class(obj), "\n")
  cat("Attempting to coerce to matrix...\n")
  counts <- as(obj, "dgCMatrix")
}

cat("Count matrix class:", class(counts), "\n")
```

---

## Per-sample exploration

Since this is a single integrated object, the checks below are applied to the full count matrix. If sample-level information is available in the metadata, we also report per-sample breakdowns.

### Matrix dimensions (`dim`)

We check the number of genes (rows) and cells (columns). The expected format is **genes × cells**.

```{r dim-counts}
cat("dim(counts):", dim(counts), "\n")
cat("  -> Genes:", nrow(counts), " | Cells:", ncol(counts), "\n")
```

### Gene identifiers (`head(rownames)`)

We inspect the first gene names to identify the type of annotation used (gene symbols, Ensembl IDs, etc.).

```{r head-rownames}
cat("head(rownames(counts)):\n")
print(head(rownames(counts)))
```

### First 20 genes (`head(rownames, 20)`)

An expanded listing of the first 20 genes for a more detailed inspection of the naming convention.

```{r head-rownames-20}
cat("head(rownames(counts), 20):\n")
print(head(rownames(counts), 20))
```

```{r gene-id-type}
# Automatic detection of gene identifier type
first_genes <- head(rownames(counts), 10)

gene_id_type <- if (any(grepl("^ENSG", first_genes))) {
  "Ensembl"
} else if (any(grepl("^[A-Z][A-Z0-9]+$", first_genes))) {
  "Gene Symbol"
} else {
  "Unknown"
}

cat("Detected gene identifier type:", gene_id_type, "\n")
```

### Cell barcodes (`head(colnames)`)

We check the format of cell barcodes (matrix columns). Typical 10x Genomics barcodes follow the pattern `ACGT...ACGT-1`.

```{r head-colnames}
cat("head(colnames(counts)):\n")
print(head(colnames(counts)))
```

### Sequencing depth (`summary(colSums)`)

The library size (total UMI count per cell) indicates sequencing depth. Very low values may indicate low-quality cells (empty droplets).

```{r summary-colsums}
cat("summary(colSums(counts)):\n")
print(summary(colSums(counts)))
```

### Genes detected per cell (`summary(colSums(counts > 0))`)

The number of genes with at least 1 UMI per cell reflects transcriptional complexity. Cells with very few detected genes may be of low quality.

```{r summary-genes-detected}
cat("summary(colSums(counts > 0)):\n")
print(summary(colSums(counts > 0)))
```

### Sequencing depth histogram (`hist(log10(colSums + 1))`)

The log10-scaled distribution allows visualizing the separation between empty droplets and real cells. A bimodal or unimodal distribution with a left tail is expected.

```{r hist-colsums, fig.height=4}
hist(
  log10(colSums(counts) + 1),
  main = paste("Library size -", params$dataset_id),
  xlab = "log10(Total UMI + 1)",
  col = "steelblue",
  border = "white",
  breaks = 50
)
```

### Mitochondrial genes (`grepl("^MT-")`)

Mitochondrial genes (`MT-` prefix) are classic indicators of cellular stress. A high fraction of mitochondrial reads (>20%) typically indicates cells undergoing apoptosis or damaged during tissue dissociation.

```{r mito-genes}
mt_count <- sum(grepl("^MT-", rownames(counts)))
cat("sum(grepl('^MT-', rownames(counts))):", mt_count, "\n")
if (mt_count > 0) {
  cat("MT genes found:\n")
  print(grep("^MT-", rownames(counts), value = TRUE))
}
```

### Ribosomal genes (`grepl("^RPL|^RPS")`)

Ribosomal genes (prefixes `RPL` — large ribosomal subunit proteins — and `RPS` — small subunit) represent a significant fraction of the transcriptome. High fractions can mask relevant biological variability and are often filtered out or regressed.

```{r ribo-genes}
ribo_count <- sum(grepl("^RPL|^RPS", rownames(counts)))
cat("sum(grepl('^RPL|^RPS', rownames(counts))):", ribo_count, "\n")
```

### Hemoglobin genes (`grepl("^HB")`)

Hemoglobin genes (e.g., `HBA1`, `HBA2`, `HBB`) indicate red blood cell contamination. Their excessive presence suggests issues with tissue dissociation or erythrocyte lysis.

```{r hb-genes}
hb_count <- sum(grepl("^HB[ABDEG]", rownames(counts)))
cat("sum(grepl('^HB', rownames(counts))):", hb_count, "\n")
if (hb_count > 0) {
  cat("HB genes found:\n")
  print(grep("^HB[ABDEG]", rownames(counts), value = TRUE))
}
```

### Duplicated genes (`sum(duplicated(rownames))`)

Duplicated gene names cause issues when creating Seurat/SingleCellExperiment objects and in downstream analyses. If present, they should be handled (e.g., `make.unique()`).

```{r duplicated-genes}
dup_count <- sum(duplicated(rownames(counts)))
cat("sum(duplicated(rownames(counts))):", dup_count, "\n")
if (dup_count > 0) {
  dups <- rownames(counts)[duplicated(rownames(counts))]
  cat("First duplicated genes:\n")
  print(head(dups, 10))
}
```

### Metadata check

Since this is a pre-built Seurat object, metadata is expected in the `meta.data` slot. We check for relevant fields such as **cell type**, **patient**, **tissue**, and **condition**.

```{r metadata-check}
if (inherits(obj, "Seurat")) {
  meta <- obj@meta.data
  cat("Metadata dimensions:", dim(meta), "\n")
  cat("  -> Cells:", nrow(meta), " | Fields:", ncol(meta), "\n\n")
  cat("Available metadata columns:\n")
  cat(paste(" ", colnames(meta)), sep = "\n")
} else {
  meta <- NULL
  cat("Object is not a Seurat object; no embedded metadata found.\n")
}
```

```{r metadata-fields}
expected_fields <- c("celltype", "cell_type", "cell.type",
                     "patient", "donor", "subject",
                     "tissue", "organ",
                     "condition", "disease", "treatment",
                     "sample", "cluster", "annotation",
                     "orig.ident", "seurat_clusters")

if (!is.null(meta)) {
  matches <- colnames(meta)[tolower(colnames(meta)) %in% expected_fields]
  cat("Fields of interest found:",
      ifelse(length(matches) > 0, paste(matches, collapse = ", "), "none"),
      "\n\n")

  # Show unique values for key metadata columns (up to 20)
  for (field in colnames(meta)) {
    if (tolower(field) %in% expected_fields) {
      cat("\n===", field, "===\n")
      vals <- unique(meta[[field]])
      cat("Unique values:", length(vals), "\n")
      if (length(vals) <= 30) {
        print(sort(vals))
      } else {
        cat("First 30 values:\n")
        print(head(sort(vals), 30))
      }
    }
  }
} else {
  cat("Metadata availability:\n")
  cat("  - celltype:  NOT AVAILABLE\n")
  cat("  - patient:   NOT AVAILABLE\n")
  cat("  - tissue:    NOT AVAILABLE\n")
  cat("  - condition: NOT AVAILABLE\n")
}
```

### Per-sample cell counts

If sample-level identifiers exist in the metadata, we report the number of cells per sample.

```{r per-sample-counts}
if (!is.null(meta)) {
  # Try common sample identifier columns
  sample_col <- NULL
  for (col in c("orig.ident", "sample", "patient", "donor", "subject")) {
    if (col %in% colnames(meta)) {
      sample_col <- col
      break
    }
  }

  if (!is.null(sample_col)) {
    cat("Using column '", sample_col, "' as sample identifier:\n\n", sep = "")
    sample_counts <- sort(table(meta[[sample_col]]), decreasing = TRUE)
    print(data.frame(
      Sample = names(sample_counts),
      Cells = as.integer(sample_counts),
      row.names = NULL
    ))
    cat("\nTotal samples:", length(sample_counts), "\n")
  } else {
    cat("No sample identifier column found in metadata.\n")
  }
} else {
  cat("No metadata available for per-sample breakdown.\n")
}
```

---

## Overall dataset summary

Summary table with the main characteristics identified in this exploration.

```{r summary-table}
mt_present   <- sum(grepl("^MT-", rownames(counts))) > 0
ribo_present <- sum(grepl("^RPL|^RPS", rownames(counts))) > 0
hb_present   <- sum(grepl("^HB[ABDEG]", rownames(counts))) > 0
dup_present  <- sum(duplicated(rownames(counts))) > 0

# Determine metadata availability
meta_available <- if (!is.null(meta)) {
  key_fields <- colnames(meta)[tolower(colnames(meta)) %in%
    c("celltype", "cell_type", "cell.type", "annotation", "cluster", "seurat_clusters")]
  ifelse(length(key_fields) > 0,
         paste("Available (", paste(key_fields, collapse = ", "), ")", sep = ""),
         "Partial (no cell type annotation)")
} else {
  "Not available"
}

# Count samples if possible
n_samples <- if (!is.null(meta) && "orig.ident" %in% colnames(meta)) {
  length(unique(meta$orig.ident))
} else if (!is.null(meta) && "sample" %in% colnames(meta)) {
  length(unique(meta$sample))
} else {
  "Unknown"
}

summary_info <- data.frame(
  Field = c(
    "Dataset",
    "Object type",
    "Samples (libraries)",
    "Total genes",
    "Total cells",
    "Gene ID type",
    "Mitochondrial genes",
    "Ribosomal genes",
    "Hemoglobin genes",
    "Duplicated genes",
    "Metadata available"
  ),
  Value = c(
    params$dataset_id,
    class(obj)[1],
    as.character(n_samples),
    nrow(counts),
    ncol(counts),
    gene_id_type,
    ifelse(mt_present, "Present", "Absent"),
    ifelse(ribo_present, "Present", "Absent"),
    ifelse(hb_present, "Present", "Absent"),
    ifelse(dup_present, "Present", "Absent"),
    meta_available
  )
)

knitr::kable(summary_info, caption = paste("Dataset summary -", params$dataset_id))
```

---

*Single-cell RNA-seq dataset initial exploration notebook.*
