---
title: "QC Exploration"
author: "Juliana Pinto"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
params:
  dataset_id: "GSExxxxxx"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# QC Exploration: `r params$dataset_id`

This notebook explores the QC metrics computed by `03_qc_metrics.R` for the `r params$dataset_id` dataset, before filtering. The aim is to choose informed thresholds for `04_qc_filtering.R`.

## Setup

```{r libs}
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)

theme_set(theme_classic())
```

```{r load-config}
config_path <- file.path(
  "~/doutorado/deconv_gastric_cancer/scripts/sc_pre_proc/configs",
  paste0("config_", params$dataset_id, ".R")
)
source(config_path)
```

```{r load-objects}
qc_files <- list.files(
  seurat_qc_metrics_dir,
  pattern = "_seurat_qc_metrics.rds$",
  full.names = TRUE
)
sample_names <- gsub("_seurat_qc_metrics.rds", "", basename(qc_files))
seu_list <- setNames(lapply(qc_files, readRDS), sample_names)

cat("Samples loaded:", length(seu_list), "\n")
cat("Names:", paste(names(seu_list), collapse = ", "), "\n")
```

## 1. Per-sample summary

```{r summary-table}
qc_summary <- do.call(rbind, lapply(names(seu_list), function(s) {
  seu <- seu_list[[s]]
  data.frame(
    sample          = s,
    cells           = ncol(seu),
    median_features = median(seu$nFeature_RNA),
    median_counts   = median(seu$nCount_RNA),
    median_mt       = round(median(seu$percent.mt), 2),
    median_ribo     = round(median(seu$percent.ribo), 2),
    median_hb       = round(median(seu$percent.hb), 2),
    pct_doublets    = if ("scDblFinder.class" %in% colnames(seu@meta.data))
      round(100 * mean(seu$scDblFinder.class == "doublet"), 1) else NA
  )
}))

knitr::kable(qc_summary, caption = paste("QC summary -", params$dataset_id))
cat("Total cells:", sum(qc_summary$cells), "\n")
```

## 2. Combined metadata

```{r combined-meta}
combined_meta <- do.call(rbind, lapply(names(seu_list), function(s) {
  df <- seu_list[[s]]@meta.data
  df$sample <- s
  df
}))
```

## 3. Comparative violin plots

```{r violins, fig.height=8}
metrics <- c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.ribo", "percent.hb")

plots <- lapply(metrics, function(m) {
  ggplot(combined_meta, aes(x = sample, y = .data[[m]], fill = sample)) +
    geom_violin(scale = "width") +
    geom_boxplot(width = 0.1, fill = "white", outlier.size = 0.3) +
    ggtitle(m) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none")
})

wrap_plots(plots, ncol = 2)
```

## 4. Distributions and candidate thresholds

```{r distributions, fig.height=8}
p1 <- ggplot(combined_meta, aes(x = nFeature_RNA, fill = sample)) +
  geom_density(alpha = 0.4) + scale_x_log10() +
  ggtitle("nFeature_RNA (log10)")

p2 <- ggplot(combined_meta, aes(x = nCount_RNA, fill = sample)) +
  geom_density(alpha = 0.4) + scale_x_log10() +
  ggtitle("nCount_RNA (log10)")

p3 <- ggplot(combined_meta, aes(x = percent.mt, fill = sample)) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = 20, linetype = "dashed", color = "red") +
  ggtitle("% mitochondrial")

p4 <- ggplot(combined_meta, aes(x = percent.ribo, fill = sample)) +
  geom_density(alpha = 0.4) +
  ggtitle("% ribosomal")

(p1 + p2) / (p3 + p4)
```

## 5. Doublets

```{r doublets}
if ("scDblFinder.class" %in% colnames(combined_meta)) {
  doublet_table <- combined_meta %>%
    group_by(sample, scDblFinder.class) %>%
    summarise(n = n(), .groups = "drop") %>%
    tidyr::pivot_wider(names_from = scDblFinder.class, values_from = n, values_fill = 0)
  knitr::kable(doublet_table, caption = "Doublet calls per sample")
}
```

## 6. Notes / decisions

- Suggested cutoffs to set in the config:
  - `min_features`:
  - `max_features`:
  - `max_percent_mt`:
- Samples needing special attention:
- Doublet removal strategy:

---

*Template — duplicate and rename to `QC_exploration_<DATASET>.qmd` per dataset, or render parametrically:*
*`quarto render 00_qc_template.qmd -P dataset_id:GSExxxxxx --output QC_exploration_GSExxxxxx.html`*
