---
title: "QC_exploration_GSE246662"
author: "Juliana Pinto"
format: html
editor: visual
---

## GSE246662

## Loading packages

```{r}
library(Seurat)
library(ggplot2)


```

Config path and .rds path

```{r}
#source("~/doutorado/deconv_gastric_cancer/scripts/sc_pre_proc/configs/config_GSE246662.R")

qc_dir <- file.path(
  project_root,
  "data",
  "sc_reference",
  "processed",
  "GSE246662",
  "seurat_qc_metrics"
)
```

Vector with path to files

```{r}
qc_files <- list.files(

  qc_dir,

  pattern = "_seurat_qc_metrics.rds$",

  full.names = TRUE

)
```

Create a list with sample names

```{r}
sample_names <- gsub("_seurat_qc_metrics.rds", "", basename(qc_files))

seu_list <- setNames(
  lapply(qc_files, readRDS),
  sample_names
)
```

Seurat objects "names"(samples in the dataset)

```{r}
names(seu_list)
```

Number of cells per sample

```{r}
sapply(seu_list, ncol)
```

```{r}
sapply(seu_list, function(x) mean(x$nFeature_RNA))
```

Plots

```{r}
for (s in names(seu_list)) {
  
  cat("##", s, "\n")
  
  seu <- seu_list[[s]]
  
  print(
    VlnPlot(
      seu,
      features = c("nFeature_RNA", "nCount_RNA"),
      pt.size = 0
    )
  )
}
```

Percentage plots

```{r}
for (s in names(seu_list)) {
  
  cat("##", s, "\n")
  
  seu <- seu_list[[s]]
  
  print(
    VlnPlot(
      seu,
      features = c("percent.mt", "percent.ribo", "percent.hb"),
      pt.size = 0
    )
  )
}

```

Histograms

```{r}
library(ggplot2)

for (s in names(seu_list)) {
  
  cat("##", s, "\n")
  
  seu <- seu_list[[s]]
  
  df <- seu@meta.data
  
  p1 <- ggplot(df, aes(x = nFeature_RNA)) +
    geom_histogram(bins = 50) +
    ggtitle(paste("nFeature_RNA -", s))
  
  print(p1)
  
  p2 <- ggplot(df, aes(x = percent.mt)) +
    geom_histogram(bins = 50) +
    ggtitle(paste("percent.mt -", s))
  
  print(p2)
}
```

a

```{r}
for (s in names(seu_list)) {
  
  cat("##", s, "\n")
  
  seu <- seu_list[[s]]
  df  <- seu@meta.data
  
  cutoff <- quantile(df$percent.mt, 0.99)
  
  p <- ggplot(df, aes(x = percent.mt)) +
    geom_density() +
    geom_vline(xintercept = cutoff, linetype = "dashed") +
    ggtitle(paste("percent.mt density -", s)) +
    theme_classic()
  
  print(p)
}
```

scatter

```{r}
 for (s in names(seu_list)) {
  
  cat("##", s, "\n")
  
  seu <- seu_list[[s]]
  df  <- seu@meta.data
  
  # -----------------------------
  # nCount_RNA vs percent.mt
  # -----------------------------
  
  p1 <- ggplot(df, aes(x = nCount_RNA, y = percent.mt)) +
    geom_point(alpha = 0.3, size = 0.5) +
    ggtitle(paste("nCount_RNA vs percent.mt -", s)) +
    theme_classic()
  
  print(p1)
  
  # --------------------------------
  # nCount_RNA vs nFeature_RNA
  # --------------------------------
  
  p2 <- ggplot(df, aes(x = nCount_RNA, y = nFeature_RNA)) +
    geom_point(alpha = 0.3, size = 0.5) +
    ggtitle(paste("nCount_RNA vs nFeature_RNA -", s)) +
    theme_classic()
  
  print(p2)
}
```

```{r}
table(seu$scDblFinder.class)
```

Doublets per sample

```{r}
for (s in names(seu_list)) {
  
  cat("\n====================\n")
  cat("Sample:", s, "\n")
  cat("====================\n")
  
  seu <- seu_list[[s]]
  
  print(table(seu$scDblFinder.class))
  
}
```

Doublet distribution

```{r}
for (s in names(seu_list)) {
  
  seu <- seu_list[[s]]
  
  print(
    VlnPlot(
      seu,
      features = "doublet_score"
    ) + ggtitle(s)
  )
  
}
```

a

```{r}
for (s in names(seu_list)) {
  
  seu <- seu_list[[s]]
  
  print(
    FeatureScatter(
      seu,
      feature1 = "nCount_RNA",
      feature2 = "doublet_score"
    ) + ggtitle(s)
  )
  
}
```

QC metrics

```{r}
library(ggplot2)

for (s in names(seu_list)) {
  
  seu <- seu_list[[s]]
  
  df <- seu@meta.data
  
  print(
    ggplot(df, aes(x = nFeature_RNA)) +
      geom_histogram(bins = 100) +
      ggtitle(paste("nFeature_RNA —", s))
  )
  
}

```

Histogram

```{r}
for (s in names(seu_list)) {
  
  df <- seu_list[[s]]@meta.data
  
  print(
    ggplot(df, aes(x = nCount_RNA)) +
      geom_histogram(bins = 100) +
      ggtitle(paste("nCount_RNA —", s))
  )
  
}
```

mt

```{r}
for (s in names(seu_list)) {
  
  df <- seu_list[[s]]@meta.data
  
  print(
    ggplot(df, aes(x = percent.mt)) +
      geom_histogram(bins = 100) +
      ggtitle(paste("percent.mt —", s))
  )
  
}
```

counts x features

```{r}
for (s in names(seu_list)) {
  
  seu <- seu_list[[s]]
  
  print(
    FeatureScatter(
      seu,
      "nCount_RNA",
      "nFeature_RNA"
    ) + ggtitle(s)
  )
  
}
```

counts x mt

```{r}
for (s in names(seu_list)) {
  
  seu <- seu_list[[s]]
  
  print(
    FeatureScatter(
      seu,
      "nCount_RNA",
      "percent.mt"
    ) + ggtitle(s)
  )
  
}
```

violin

```{r}
for (s in names(seu_list)) {
  
  df <- seu_list[[s]]@meta.data
  
  cor_matrix <- cor(
    df[, c("nFeature_RNA","nCount_RNA","percent.mt")],
    method = "spearman"
  )
  
  cat("\nCorrelation matrix —", s, "\n")
  print(cor_matrix)
  
}
```

s

```{r}
qc_summary <- do.call(
  rbind,
  lapply(names(seu_list), function(s){
    
    seu <- seu_list[[s]]
    
    data.frame(
      sample = s,
      cells = ncol(seu),
      mean_features = mean(seu$nFeature_RNA),
      mean_counts = mean(seu$nCount_RNA),
      mean_mt = mean(seu$percent.mt)
    )
    
  })
)

qc_summary
```

-   A amostra tal nao serguiraá analises, por alto %MT
