---
title: "Pre-Annotation Exploration - GSE163558"
author: "Juliana Pinto"
format: html
editor: visual
---

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(ggplot2)
library(patchwork)
library(pheatmap)
library(dplyr)

theme_set(theme_classic())
```

```{r load-config}
source("~/doutorado/deconv_gastric_cancer/scripts/sc_pre_proc/configs/config_GSE163558.R")
```

```{r helper-function}
ens_to_label <- function(ensembl_ids, seu) {
  if ("gene_symbol" %in% colnames(seu[["RNA"]]@meta.data)) {
    gene_meta <- seu[["RNA"]]@meta.data
    symbol_lookup <- setNames(gene_meta$gene_symbol, rownames(seu))
    labels <- symbol_lookup[ensembl_ids]
    labels[is.na(labels) | labels == ""] <- ensembl_ids[is.na(labels) | labels == ""]
    return(unname(labels))
  }
  return(ensembl_ids)
}
```

## 1. Load annotated object

```{r load-object}
seu <- readRDS(annotated_seurat_path)

cat("Dataset:", dataset_id, "\n")
cat("Dimensions:", nrow(seu), "genes x", ncol(seu), "cells\n")
cat("Samples:", length(unique(seu$orig.ident)), "\n")

# Check which annotations are available
annot_cols <- grep("singler|celltypist", colnames(seu@meta.data), value = TRUE)
cat("Annotation columns:", paste(annot_cols, collapse = ", "), "\n")
```

## 2. UMAP - All annotations

```{r umap-all, fig.width=20, fig.height=16}
annotation_cluster_col <- paste0("RNA_snn_res.", annotation_resolution)

p1 <- DimPlot(seu, group.by = annotation_cluster_col, label = TRUE) +
  ggtitle(paste("Clusters (res", annotation_resolution, ")")) + NoLegend()

plots <- list(p1)

if ("singler_hpca" %in% colnames(seu@meta.data)) {
  plots[["hpca"]] <- DimPlot(seu, group.by = "singler_hpca", label = TRUE, repel = TRUE) +
    ggtitle("SingleR - HPCA") + NoLegend()
}

if ("singler_blueprint" %in% colnames(seu@meta.data)) {
  plots[["bp"]] <- DimPlot(seu, group.by = "singler_blueprint", label = TRUE, repel = TRUE) +
    ggtitle("SingleR - Blueprint") + NoLegend()
}

if ("celltypist_label" %in% colnames(seu@meta.data)) {
  plots[["ct"]] <- DimPlot(seu, group.by = "celltypist_label", label = TRUE, repel = TRUE) +
    ggtitle("CellTypist") + NoLegend()
}

wrap_plots(plots, ncol = 2)
```

## 3. Annotation counts

```{r annotation-counts}
if ("singler_hpca" %in% colnames(seu@meta.data)) {
  cat("--- SingleR HPCA ---\n")
  print(sort(table(seu$singler_hpca), decreasing = TRUE))
  cat("\n")
}

if ("singler_blueprint" %in% colnames(seu@meta.data)) {
  cat("--- SingleR Blueprint ---\n")
  print(sort(table(seu$singler_blueprint), decreasing = TRUE))
  cat("\n")
}

if ("celltypist_label" %in% colnames(seu@meta.data)) {
  cat("--- CellTypist ---\n")
  print(sort(table(seu$celltypist_label), decreasing = TRUE))
}
```

## 4. CellTypist confidence scores

```{r celltypist-confidence, fig.width=10, fig.height=10}
if ("celltypist_conf_score" %in% colnames(seu@meta.data)) {
  p1 <- FeaturePlot(seu, features = "celltypist_conf_score", order = TRUE) +
    scale_color_gradient(low = "red", high = "steelblue", name = "Score") +
    ggtitle("CellTypist confidence on UMAP")

  p2 <- ggplot(seu@meta.data, aes(x = celltypist_conf_score)) +
    geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
    geom_vline(xintercept = 0.5, linetype = "dashed", color = "red") +
    ggtitle("CellTypist confidence distribution") +
    xlab("Confidence score") + ylab("Cells")

  p1 / p2
}
```

```{r low-confidence}
if ("celltypist_conf_score" %in% colnames(seu@meta.data)) {
  low_conf <- sum(seu$celltypist_conf_score < 0.5, na.rm = TRUE)
  cat("Cells with confidence < 0.5:", low_conf, "(", round(100 * low_conf / ncol(seu), 1), "%)\n")

  # Which labels have lowest confidence?
  conf_by_label <- seu@meta.data %>%
    group_by(celltypist_label) %>%
    summarise(
      n = n(),
      mean_conf = round(mean(celltypist_conf_score, na.rm = TRUE), 3),
      pct_low = round(sum(celltypist_conf_score < 0.5, na.rm = TRUE) / n() * 100, 1),
      .groups = "drop"
    ) %>%
    arrange(mean_conf)

  knitr::kable(conf_by_label, caption = "CellTypist confidence by label")
}
```

## 5. Cluster-level consensus

```{r cluster-consensus}
annotation_cluster_col <- paste0("RNA_snn_res.", annotation_resolution)

consensus <- seu@meta.data %>%
  group_by(across(all_of(annotation_cluster_col))) %>%
  summarise(
    n_cells = n(),
    hpca = if ("singler_hpca" %in% colnames(seu@meta.data))
      names(sort(table(singler_hpca), decreasing = TRUE))[1] else NA,
    hpca_pct = if ("singler_hpca" %in% colnames(seu@meta.data))
      round(max(table(singler_hpca)) / n() * 100, 1) else NA,
    blueprint = if ("singler_blueprint" %in% colnames(seu@meta.data))
      names(sort(table(singler_blueprint), decreasing = TRUE))[1] else NA,
    blueprint_pct = if ("singler_blueprint" %in% colnames(seu@meta.data))
      round(max(table(singler_blueprint)) / n() * 100, 1) else NA,
    celltypist = if ("celltypist_label" %in% colnames(seu@meta.data))
      names(sort(table(celltypist_label), decreasing = TRUE))[1] else NA,
    celltypist_pct = if ("celltypist_label" %in% colnames(seu@meta.data))
      round(max(table(celltypist_label)) / n() * 100, 1) else NA,
    .groups = "drop"
  )

knitr::kable(consensus, caption = "Dominant annotation per cluster (% agreement)")
```

## 6. Confusion matrices

### SingleR HPCA vs Clusters

```{r confusion-hpca, fig.width=14, fig.height=8}
if ("singler_hpca" %in% colnames(seu@meta.data)) {
  ct <- table(
    Cluster = seu@meta.data[[annotation_cluster_col]],
    HPCA = seu$singler_hpca
  )
  ct_prop <- prop.table(ct, margin = 1) * 100

  pheatmap(
    ct_prop,
    fontsize = 7,
    display_numbers = TRUE,
    number_format = "%.0f",
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    color = colorRampPalette(c("white", "steelblue"))(100),
    main = "Cluster vs SingleR-HPCA (%)"
  )
}
```

### SingleR Blueprint vs Clusters

```{r confusion-blueprint, fig.width=14, fig.height=8}
if ("singler_blueprint" %in% colnames(seu@meta.data)) {
  ct <- table(
    Cluster = seu@meta.data[[annotation_cluster_col]],
    Blueprint = seu$singler_blueprint
  )
  ct_prop <- prop.table(ct, margin = 1) * 100

  pheatmap(
    ct_prop,
    fontsize = 7,
    display_numbers = TRUE,
    number_format = "%.0f",
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    color = colorRampPalette(c("white", "steelblue"))(100),
    main = "Cluster vs SingleR-Blueprint (%)"
  )
}
```

### CellTypist vs SingleR HPCA

```{r confusion-ct-hpca, fig.width=16, fig.height=10}
if (all(c("celltypist_label", "singler_hpca") %in% colnames(seu@meta.data))) {
  ct <- table(
    CellTypist = seu$celltypist_label,
    HPCA = seu$singler_hpca
  )
  ct_prop <- prop.table(ct, margin = 1) * 100

  pheatmap(
    ct_prop,
    fontsize = 6,
    display_numbers = TRUE,
    number_format = "%.0f",
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    color = colorRampPalette(c("white", "steelblue"))(100),
    main = "CellTypist vs SingleR-HPCA (%)"
  )
}
```

### CellTypist vs SingleR Blueprint

```{r confusion-ct-bp, fig.width=16, fig.height=10}
if (all(c("celltypist_label", "singler_blueprint") %in% colnames(seu@meta.data))) {
  ct <- table(
    CellTypist = seu$celltypist_label,
    Blueprint = seu$singler_blueprint
  )
  ct_prop <- prop.table(ct, margin = 1) * 100

  pheatmap(
    ct_prop,
    fontsize = 6,
    display_numbers = TRUE,
    number_format = "%.0f",
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    color = colorRampPalette(c("white", "steelblue"))(100),
    main = "CellTypist vs SingleR-Blueprint (%)"
  )
}
```

## 7. By sample

```{r annotation-by-sample, fig.width=14, fig.height=6}
if ("singler_hpca" %in% colnames(seu@meta.data)) {
  df <- seu@meta.data %>%
    group_by(orig.ident, singler_hpca) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(orig.ident) %>%
    mutate(pct = n / sum(n) * 100)

  ggplot(df, aes(x = orig.ident, y = pct, fill = singler_hpca)) +
    geom_col() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggtitle("Cell type composition per sample (SingleR-HPCA)") +
    ylab("% cells") + xlab(NULL)
}
```

## 8. Summary

```{r summary}
cat("Dataset:", dataset_id, "\n")
cat("Cells:", ncol(seu), "\n")
cat("Resolution:", annotation_resolution, "\n")
cat("Annotations available:", paste(annot_cols, collapse = ", "), "\n")

if ("singler_hpca" %in% colnames(seu@meta.data))
  cat("SingleR HPCA labels:", length(unique(seu$singler_hpca)), "\n")
if ("singler_blueprint" %in% colnames(seu@meta.data))
  cat("SingleR Blueprint labels:", length(unique(seu$singler_blueprint)), "\n")
if ("celltypist_label" %in% colnames(seu@meta.data)) {
  cat("CellTypist labels:", length(unique(na.omit(seu$celltypist_label))), "\n")
  cat("CellTypist mean confidence:", round(mean(seu$celltypist_conf_score, na.rm = TRUE), 3), "\n")
}
```

## 9. Conclusions

Notes:

-   **Agreement**: Where do all three methods agree? Those are high-confidence cell type assignments.
-   **Disagreements**: Where they disagree, which method seems more plausible given the known markers?
-   **CellTypist confidence**: Cells with \< 0.5 confidence may be misclassified or represent transitional states.
-   **Batch effect**: Do annotations vary by sample? This indicates batch effect still present (expected pre-integration).
-   **Epithelial cells**: CellTypist (immune-focused model) may struggle with epithelial cells. SingleR with HPCA is better for these.
-   **Reminder**: This is pre-annotation. Final annotation will be done on the integrated object.
